---
title: "The Molecular Centaur — Complete Analysis"
subtitle: "All validations and tests, Sections 1–9"
author: "Daniel E. Yehoshua"
date: "`r Sys.Date()`"
params:
  repository_url: "https://github.com/USERNAME/REPOSITORY"
  install_missing_packages: false
  download_tcga_if_missing: true
  build_xcell2_if_missing: true
  download_gse176078_if_missing: true
  build_gse176078_seurat_if_missing: true
  gse176078_external_dir: "data/external/GSE176078"
  seurat_object_path: "data/GSE176078_seurat.rds"
  run_foxc1_motif_test: false
  run_scrna_ccl19_test: false
  foxc1_motif_min_score: "85%"
  promoter_upstream_bp: 2000
  promoter_downstream_bp: 200
  motif_background_genes: 500
  force_rebuild_cache: false
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_caption: true
    df_print: kable
---

<style>
.main-container { max-width: 1200px; }
h1 { border-bottom: 2px solid #2c3e50; padding-bottom: 8px; margin-top: 40px; }
h2 { color: #2c3e50; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 11, fig.height = 7, fig.align = "center",
  dpi = 150, out.width = "100%")
options(knitr.kable.NA = "")

# This report is designed to be self-contained for repository use:
# cached inputs are loaded when present; otherwise public TCGA-derived inputs
# are downloaded/rebuilt when params$download_tcga_if_missing is TRUE.
```



# Setup and Data Construction

# Repository and Reproducibility

This report is intended to be the executable analysis record for the manuscript.

```{r repository_metadata}
repo_url <- params$repository_url
cat("Repository:", repo_url, "\n")
cat("Generated:", as.character(Sys.time()), "\n")
cat("R version:", R.version.string, "\n")
```

**Reproducibility policy.** The report first looks for cached RDS objects under `results/rds/`. If the required files are missing and `params$download_tcga_if_missing = TRUE`, it downloads and rebuilds the TCGA-BRCA expression, subtype, and clinical inputs. Expensive derived objects, including xCell2 scores, are cached after creation. The single-cell validation object path is controlled by `params$seurat_object_path`; if absent and `params$build_gse176078_seurat_if_missing = TRUE`, the report downloads GSE176078 supplementary files, reconstructs a Seurat object when a supported processed format is detected, standardizes the published annotation column into `celltype_subset`, normalizes the RNA assay if needed, and saves the result to `params$seurat_object_path`. Heavy exploratory mechanism tests (FOXC1 promoter motif scanning and single-cell FOXC1/CCL19 co-expression by cell type) are included but disabled by default; enable them with `params$run_foxc1_motif_test = TRUE` and/or `params$run_scrna_ccl19_test = TRUE` when the required dependencies and inputs are available.

```{r package_bootstrap, include=FALSE}
ensure_packages <- function(pkgs, bioc = FALSE) {
  missing <- pkgs[!vapply(pkgs, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing) == 0) return(invisible(TRUE))
  if (!isTRUE(params$install_missing_packages)) {
    stop(
      "Missing required package(s): ", paste(missing, collapse = ", "),
      "\nInstall them manually, or knit with params$install_missing_packages = TRUE.",
      call. = FALSE
    )
  }
  if (bioc) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
    BiocManager::install(missing, ask = FALSE, update = FALSE)
  } else {
    install.packages(missing, repos = "https://cloud.r-project.org")
  }
  invisible(TRUE)
}

ensure_packages(c("tidyverse", "survival", "survminer", "pheatmap", "patchwork"), bioc = FALSE)
ensure_packages(c("AnnotationDbi", "org.Hs.eg.db"), bioc = TRUE)
```

```{r libraries}
library(tidyverse)
library(survival)
library(survminer)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(pheatmap)
library(patchwork)
suppressWarnings({
  has_gsva       <- requireNamespace("GSVA", quietly = TRUE)
  has_diptest    <- requireNamespace("diptest", quietly = TRUE)
  has_mclust     <- requireNamespace("mclust", quietly = TRUE)
  has_seurat     <- requireNamespace("Seurat", quietly = TRUE)
  has_xcell2     <- requireNamespace("xCell2", quietly = TRUE)
  has_tcgabiolinks <- requireNamespace("TCGAbiolinks", quietly = TRUE)
})
```

```{r reproducible_input_cache}
# ---- Directory layout used by the Git repository ----
dir.create("data", recursive = TRUE, showWarnings = FALSE)
dir.create("data/external", recursive = TRUE, showWarnings = FALSE)
dir.create("data/processed", recursive = TRUE, showWarnings = FALSE)
dir.create("results", recursive = TRUE, showWarnings = FALSE)
dir.create("results/rds", recursive = TRUE, showWarnings = FALSE)
dir.create("results/tables", recursive = TRUE, showWarnings = FALSE)
dir.create("results/figures", recursive = TRUE, showWarnings = FALSE)

core_rds <- c(
  norm_counts = "results/rds/norm_counts.rds",
  sample_meta = "results/rds/sample_meta.rds",
  brca_clin   = "results/rds/brca_clin.rds"
)

optional_rds <- c(
  full_counts      = "results/rds/full_counts.rds",
  tpm_counts_tumor = "results/rds/tpm_counts_tumor.rds",
  xcell2_scores    = "results/rds/xcell2_scores.rds"
)

standardize_tcga_barcode <- function(x, n = NULL) {
  y <- toupper(gsub("\\.", "-", as.character(x)))
  if (!is.null(n)) y <- substr(y, 1, n)
  y
}

pick_first_existing <- function(x, candidates) {
  hit <- candidates[candidates %in% names(x)]
  if (length(hit) == 0) return(NA_character_)
  hit[1]
}

build_tcga_brca_cache <- function() {
  needed <- c("TCGAbiolinks", "SummarizedExperiment", "DESeq2")
  missing <- needed[!vapply(needed, requireNamespace, logical(1), quietly = TRUE)]
  if (length(missing) > 0) {
    if (isTRUE(params$install_missing_packages)) {
      ensure_packages(missing, bioc = TRUE)
    } else {
      stop(
        "TCGA cache is missing and these packages are required to rebuild it: ",
        paste(missing, collapse = ", "),
        "\nInstall them manually, or knit with params$install_missing_packages = TRUE.",
        call. = FALSE
      )
    }
  }

  message("Downloading/rebuilding TCGA-BRCA STAR-count cache...")
  query <- TCGAbiolinks::GDCquery(
    project = "TCGA-BRCA",
    data.category = "Transcriptome Profiling",
    data.type = "Gene Expression Quantification",
    workflow.type = "STAR - Counts",
    sample.type = "Primary Tumor"
  )
  TCGAbiolinks::GDCdownload(query, method = "api", files.per.chunk = 50)
  se <- TCGAbiolinks::GDCprepare(query)

  assays_available <- names(SummarizedExperiment::assays(se))
  count_assay <- intersect(c("unstranded", "raw_count", "counts"), assays_available)[1]
  if (is.na(count_assay)) count_assay <- assays_available[1]
  full_counts <- as.matrix(SummarizedExperiment::assay(se, count_assay))

  # Defensive tumor filter; GDCquery already asked for Primary Tumor, but keep this explicit.
  sample_barcodes <- colnames(full_counts)
  tumor_keep <- substr(standardize_tcga_barcode(sample_barcodes), 14, 15) %in% c("01")
  full_counts <- full_counts[, tumor_keep, drop = FALSE]
  sample_barcodes <- colnames(full_counts)
  storage.mode(full_counts) <- "integer"
  saveRDS(full_counts, optional_rds[["full_counts"]])

  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = round(full_counts),
    colData = data.frame(row.names = sample_barcodes),
    design = ~ 1
  )
  dds <- DESeq2::estimateSizeFactors(dds)
  norm_counts <- DESeq2::counts(dds, normalized = TRUE)
  saveRDS(norm_counts, core_rds[["norm_counts"]])

  # TPM matrix for xCell2. Prefer GDC-provided TPM when available.
  tpm_assay <- intersect(c("tpm_unstranded", "TPM", "tpm"), assays_available)[1]
  if (!is.na(tpm_assay)) {
    tpm_counts_tumor <- as.matrix(SummarizedExperiment::assay(se, tpm_assay))[, sample_barcodes, drop = FALSE]
  } else {
    rd <- as.data.frame(SummarizedExperiment::rowData(se))
    len_col <- intersect(c("gene_length", "length", "transcript_length"), names(rd))[1]
    if (is.na(len_col)) {
      warning("No TPM assay or gene-length column found; xCell2 rebuild will require a TPM matrix.")
      tpm_counts_tumor <- NULL
    } else {
      gene_length_kb <- as.numeric(rd[[len_col]]) / 1000
      gene_length_kb[!is.finite(gene_length_kb) | gene_length_kb <= 0] <- NA
      rate <- full_counts / gene_length_kb
      tpm_counts_tumor <- t(t(rate) / colSums(rate, na.rm = TRUE)) * 1e6
      tpm_counts_tumor[!is.finite(tpm_counts_tumor)] <- 0
    }
  }
  if (!is.null(tpm_counts_tumor)) saveRDS(tpm_counts_tumor, optional_rds[["tpm_counts_tumor"]])

  # PAM50 subtype metadata.
  subtype <- TCGAbiolinks::TCGAquery_subtype(tumor = "BRCA")
  subtype <- as.data.frame(subtype)
  patient_col <- pick_first_existing(subtype, c("patient", "Patient", "bcr_patient_barcode", "TCGA_patient_barcode", "barcode"))
  if (is.na(patient_col)) {
    patient_col <- names(subtype)[grepl("patient|barcode|submitter", names(subtype), ignore.case = TRUE)][1]
  }
  pam50_col <- pick_first_existing(subtype, c("BRCA_Subtype_PAM50", "PAM50", "Subtype_Selected", "subtype"))
  if (is.na(pam50_col)) {
    pam50_col <- names(subtype)[grepl("PAM50|Subtype", names(subtype), ignore.case = TRUE)][1]
  }
  if (is.na(patient_col) || is.na(pam50_col)) {
    stop("Could not identify patient/PAM50 columns in TCGAquery_subtype('BRCA') output.", call. = FALSE)
  }
  subtype$patient <- standardize_tcga_barcode(subtype[[patient_col]], 12)
  subtype$pam50_from_tcga <- as.character(subtype[[pam50_col]])
  subtype <- subtype[!duplicated(subtype$patient), c("patient", "pam50_from_tcga")]

  sample_meta_tumor <- data.frame(
    barcode = sample_barcodes,
    patient = standardize_tcga_barcode(sample_barcodes, 12),
    stringsAsFactors = FALSE
  )
  sample_meta_tumor <- dplyr::left_join(sample_meta_tumor, subtype, by = "patient")
  sample_meta_tumor$paper_BRCA_Subtype_PAM50 <- sample_meta_tumor$pam50_from_tcga
  sample_meta_tumor$pam50_from_tcga <- NULL
  saveRDS(sample_meta_tumor, core_rds[["sample_meta"]])

  # Clinical metadata. Radiation field names have changed across GDC/TCGAbiolinks versions,
  # so this block normalizes the most common variants into the field expected downstream.
  brca.clin <- TCGAbiolinks::GDCquery_clinic(project = "TCGA-BRCA", type = "clinical")
  brca.clin <- as.data.frame(brca.clin)
  if (!"submitter_id" %in% names(brca.clin)) {
    id_col <- pick_first_existing(brca.clin, c("bcr_patient_barcode", "case_submitter_id", "patient"))
    if (!is.na(id_col)) brca.clin$submitter_id <- standardize_tcga_barcode(brca.clin[[id_col]], 12)
  }
  rad_col <- pick_first_existing(brca.clin, c(
    "treatments_radiation_treatment_or_therapy",
    "radiation_treatment_or_therapy",
    "radiation_therapy",
    "treatment_or_therapy"
  ))
  if (!is.na(rad_col)) {
    brca.clin$treatments_radiation_treatment_or_therapy <- tolower(as.character(brca.clin[[rad_col]]))
  } else {
    warning("No radiation-treatment column detected in downloaded clinical table; radiation-stratified survival will require a cached brca_clin.rds with this field.")
    brca.clin$treatments_radiation_treatment_or_therapy <- NA_character_
  }
  saveRDS(brca.clin, core_rds[["brca_clin"]])

  message("TCGA cache complete:")
  message("  norm_counts: ", core_rds[["norm_counts"]])
  message("  sample_meta: ", core_rds[["sample_meta"]])
  message("  brca_clin:   ", core_rds[["brca_clin"]])
  invisible(TRUE)
}

build_xcell2_cache <- function() {
  xf <- optional_rds[["xcell2_scores"]]
  if (file.exists(xf) && !isTRUE(params$force_rebuild_cache)) {
    message("Using cached xCell2 scores: ", xf)
    return(invisible(TRUE))
  }
  if (!isTRUE(params$build_xcell2_if_missing)) {
    message("xCell2 scores missing, but params$build_xcell2_if_missing is FALSE; skipping.")
    return(invisible(FALSE))
  }
  if (!requireNamespace("xCell2", quietly = TRUE)) {
    warning("Package xCell2 is not installed; skipping xCell2 rebuild. Install xCell2 and re-knit, or provide results/rds/xcell2_scores.rds.")
    return(invisible(FALSE))
  }
  if (!file.exists(optional_rds[["tpm_counts_tumor"]])) {
    warning("TPM matrix not found at ", optional_rds[["tpm_counts_tumor"]], "; cannot rebuild xCell2 scores.")
    return(invisible(FALSE))
  }
  if (!requireNamespace("BiocParallel", quietly = TRUE)) {
    if (isTRUE(params$install_missing_packages)) ensure_packages("BiocParallel", bioc = TRUE)
    else warning("BiocParallel is missing; xCell2 will try to run without explicit SerialParam.")
  }

  message("Building xCell2 scores from TPM matrix...")
  tpm_counts_tumor <- readRDS(optional_rds[["tpm_counts_tumor"]])
  ens <- gsub("\\..*", "", rownames(tpm_counts_tumor))
  gene_symbols <- AnnotationDbi::mapIds(
    org.Hs.eg.db,
    keys = ens,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first"
  )
  tpm_sym <- tpm_counts_tumor[!is.na(gene_symbols), , drop = FALSE]
  rownames(tpm_sym) <- gene_symbols[!is.na(gene_symbols)]
  tpm_sym <- rowsum(tpm_sym, group = rownames(tpm_sym), reorder = FALSE)

  data("PanCancer.xCell2Ref", package = "xCell2")
  bp <- if (requireNamespace("BiocParallel", quietly = TRUE)) BiocParallel::SerialParam() else NULL
  xcell2_scores <- tryCatch({
    if (is.null(bp)) {
      xCell2::xCell2Analysis(mix = tpm_sym, xcell2object = PanCancer.xCell2Ref, spillover = TRUE)
    } else {
      xCell2::xCell2Analysis(mix = tpm_sym, xcell2object = PanCancer.xCell2Ref, spillover = TRUE, BPPARAM = bp)
    }
  }, error = function(e) {
    message("xCell2 call with PanCancer.xCell2Ref failed; trying package defaults. Error was: ", conditionMessage(e))
    xCell2::xCell2Analysis(mix = tpm_sym)
  })
  saveRDS(xcell2_scores, xf)
  message("Saved xCell2 scores: ", xf)
  invisible(TRUE)
}

gse176078_dir <- function() {
  x <- tryCatch(params$gse176078_external_dir, error = function(e) NULL)
  if (is.null(x) || length(x) == 0 || is.na(x) || !nzchar(x)) "data/external/GSE176078" else x
}

fetch_gse176078_supplementary <- function() {
  sf <- params$seurat_object_path
  if (file.exists(sf) && !isTRUE(params$force_rebuild_cache)) return(invisible(TRUE))

  gse_dir <- gse176078_dir()
  if (dir.exists(gse_dir) && length(list.files(gse_dir, recursive = TRUE, all.files = FALSE)) > 0) {
    message("Using existing GSE176078 supplementary directory: ", gse_dir)
    return(invisible(TRUE))
  }

  if (!isTRUE(params$download_gse176078_if_missing)) {
    message("Single-cell object not found at ", sf,
            "; skipping GEO download because params$download_gse176078_if_missing is FALSE.")
    return(invisible(FALSE))
  }
  if (!requireNamespace("GEOquery", quietly = TRUE)) {
    if (isTRUE(params$install_missing_packages)) ensure_packages("GEOquery", bioc = TRUE)
    else stop("GEOquery is required to download GSE176078 supplementary files.", call. = FALSE)
  }

  dir.create("data/external", recursive = TRUE, showWarnings = FALSE)
  message("Downloading GSE176078 supplementary files from GEO...")
  GEOquery::getGEOSuppFiles("GSE176078", baseDir = "data/external", makeDirectory = TRUE)
  if (!dir.exists(gse_dir)) {
    # GEOquery normally creates data/external/GSE176078; keep this defensive.
    hits <- list.dirs("data/external", recursive = FALSE, full.names = TRUE)
    hit <- hits[grepl("GSE176078", basename(hits), ignore.case = TRUE)][1]
    if (!is.na(hit) && dir.exists(hit)) gse_dir <- hit
  }
  message("Downloaded GSE176078 supplementary files under: ", gse_dir)
  invisible(dir.exists(gse_dir))
}

decompress_gse176078_files <- function(gse_dir) {
  if (!dir.exists(gse_dir)) return(character())
  files <- list.files(gse_dir, recursive = TRUE, full.names = TRUE)

  # Expand archives, but leave simple .gz files in place because Read10X/readr/fread can read them.
  archive_files <- files[grepl("\\.(tar|tar\\.gz|tgz|zip)$", files, ignore.case = TRUE)]
  for (ff in archive_files) {
    marker <- file.path(dirname(ff), paste0(".extracted_", basename(ff), ".done"))
    if (file.exists(marker) && !isTRUE(params$force_rebuild_cache)) next
    message("Extracting ", ff)
    ok <- tryCatch({
      if (grepl("\\.zip$", ff, ignore.case = TRUE)) {
        utils::unzip(ff, exdir = dirname(ff))
      } else {
        utils::untar(ff, exdir = dirname(ff))
      }
      TRUE
    }, error = function(e) {
      warning("Could not extract ", ff, ": ", conditionMessage(e))
      FALSE
    })
    if (isTRUE(ok)) writeLines(as.character(Sys.time()), marker)
  }
  list.files(gse_dir, recursive = TRUE, full.names = TRUE)
}

read_table_flexible <- function(path, n_max = Inf) {
  if (requireNamespace("data.table", quietly = TRUE)) {
    if (is.infinite(n_max)) {
      return(as.data.frame(data.table::fread(path, nThread = 1, data.table = FALSE)))
    }
    return(as.data.frame(data.table::fread(path, nThread = 1, nrows = n_max, data.table = FALSE)))
  }
  sep <- if (grepl("\\.csv(\\.gz)?$", path, ignore.case = TRUE)) "," else "\t"
  if (is.infinite(n_max)) {
    return(utils::read.table(path, header = TRUE, sep = sep, quote = "\"", comment.char = "",
                             check.names = FALSE, stringsAsFactors = FALSE))
  }
  utils::read.table(path, header = TRUE, sep = sep, quote = "\"", comment.char = "",
                    check.names = FALSE, stringsAsFactors = FALSE, nrows = n_max)
}

find_seurat_object_in_serialized_files <- function(files) {
  rds_files <- files[grepl("\\.(rds|RDS)$", files)]
  for (ff in rds_files) {
    obj <- tryCatch(readRDS(ff), error = function(e) NULL)
    if (inherits(obj, "Seurat")) {
      message("Found processed Seurat object in: ", ff)
      return(obj)
    }
  }

  rdata_files <- files[grepl("\\.(rda|RData)$", files, ignore.case = TRUE)]
  for (ff in rdata_files) {
    env <- new.env(parent = emptyenv())
    ok <- tryCatch({ load(ff, envir = env); TRUE }, error = function(e) FALSE)
    if (!ok) next
    for (nm in ls(env)) {
      obj <- get(nm, envir = env)
      if (inherits(obj, "Seurat")) {
        message("Found processed Seurat object in: ", ff, " object: ", nm)
        return(obj)
      }
    }
  }
  NULL
}

load_h5ad_as_seurat <- function(files) {
  h5ad <- files[grepl("\\.h5ad$", files, ignore.case = TRUE)][1]
  if (is.na(h5ad)) return(NULL)
  if (!requireNamespace("SeuratDisk", quietly = TRUE)) {
    message("Found h5ad file but SeuratDisk is not installed; skipping h5ad import: ", h5ad)
    return(NULL)
  }
  h5s <- sub("\\.h5ad$", ".h5seurat", h5ad, ignore.case = TRUE)
  if (!file.exists(h5s) || isTRUE(params$force_rebuild_cache)) {
    message("Converting h5ad to h5Seurat: ", h5ad)
    SeuratDisk::Convert(h5ad, dest = "h5seurat", overwrite = TRUE)
  }
  message("Loading h5Seurat object: ", h5s)
  SeuratDisk::LoadH5Seurat(h5s, assays = "RNA", reductions = NULL, graphs = FALSE, images = FALSE)
}

load_10x_h5_as_seurat <- function(files) {
  h5_files <- files[grepl("\\.(h5|hdf5)$", files, ignore.case = TRUE)]
  if (length(h5_files) == 0) return(NULL)
  # Avoid h5ad here; handled above.
  h5_files <- h5_files[!grepl("\\.h5ad$", h5_files, ignore.case = TRUE)]
  if (length(h5_files) == 0) return(NULL)

  objs <- list()
  for (ff in h5_files) {
    message("Trying 10X HDF5 import: ", ff)
    mat <- tryCatch(Seurat::Read10X_h5(ff), error = function(e) NULL)
    if (is.null(mat)) next
    if (is.list(mat)) {
      nm <- intersect(c("Gene Expression", "RNA"), names(mat))[1]
      mat <- if (!is.na(nm)) mat[[nm]] else mat[[1]]
    }
    sample_id <- tools::file_path_sans_ext(basename(ff))
    colnames(mat) <- paste(sample_id, colnames(mat), sep = "_")
    objs[[sample_id]] <- Seurat::CreateSeuratObject(counts = mat, project = sample_id)
  }
  if (length(objs) == 0) return(NULL)
  if (length(objs) == 1) return(objs[[1]])
  Reduce(function(a, b) merge(a, b), objs)
}

find_10x_dirs <- function(files) {
  mtx <- files[grepl("matrix\\.mtx(\\.gz)?$", basename(files), ignore.case = TRUE)]
  dirs <- unique(dirname(mtx))
  dirs[vapply(dirs, function(d) {
    local <- basename(list.files(d, full.names = FALSE))
    any(grepl("barcodes\\.tsv(\\.gz)?$", local, ignore.case = TRUE)) &&
      any(grepl("(features|genes)\\.tsv(\\.gz)?$", local, ignore.case = TRUE))
  }, logical(1))]
}

load_10x_dirs_as_seurat <- function(files) {
  dirs <- find_10x_dirs(files)
  if (length(dirs) == 0) return(NULL)
  objs <- list()
  for (dd in dirs) {
    sample_id <- basename(dd)
    message("Trying 10X directory import: ", dd)
    mat <- tryCatch(Seurat::Read10X(data.dir = dd), error = function(e) NULL)
    if (is.null(mat)) next
    if (is.list(mat)) {
      nm <- intersect(c("Gene Expression", "RNA"), names(mat))[1]
      mat <- if (!is.na(nm)) mat[[nm]] else mat[[1]]
    }
    colnames(mat) <- paste(sample_id, colnames(mat), sep = "_")
    objs[[sample_id]] <- Seurat::CreateSeuratObject(counts = mat, project = sample_id)
  }
  if (length(objs) == 0) return(NULL)
  if (length(objs) == 1) return(objs[[1]])
  Reduce(function(a, b) merge(a, b), objs)
}

load_delimited_expression_as_seurat <- function(files) {
  expr_files <- files[grepl("(count|counts|matrix|expression|expr|data).*\\.(csv|tsv|txt)(\\.gz)?$", basename(files), ignore.case = TRUE)]
  expr_files <- expr_files[!grepl("(meta|annot|clinical|barcode|feature|gene)", basename(expr_files), ignore.case = TRUE)]
  if (length(expr_files) == 0) return(NULL)

  # Try the most likely files first, but do not read all huge files if one succeeds.
  expr_files <- expr_files[order(nchar(basename(expr_files)))]
  for (ff in expr_files) {
    message("Trying delimited expression import: ", ff)
    df <- tryCatch(read_table_flexible(ff), error = function(e) NULL)
    if (is.null(df) || ncol(df) < 3 || nrow(df) < 100) next
    gene_col <- names(df)[1]
    genes <- as.character(df[[gene_col]])
    if (anyDuplicated(genes)) genes <- make.unique(genes)
    mat_df <- df[, -1, drop = FALSE]
    mat <- tryCatch(as.matrix(mat_df), error = function(e) NULL)
    if (is.null(mat)) next
    suppressWarnings(storage.mode(mat) <- "numeric")
    if (sum(is.finite(mat), na.rm = TRUE) == 0) next
    rownames(mat) <- genes
    mat[!is.finite(mat)] <- 0
    return(Seurat::CreateSeuratObject(counts = mat, project = "GSE176078"))
  }
  NULL
}

add_gse176078_metadata <- function(seu, files) {
  md_files <- files[grepl("(meta|metadata|annot|annotation|cell).*\\.(csv|tsv|txt)(\\.gz)?$", basename(files), ignore.case = TRUE)]
  md_files <- md_files[!grepl("(matrix|count|expression)", basename(md_files), ignore.case = TRUE)]
  if (length(md_files) == 0) return(seu)

  for (ff in md_files) {
    message("Trying metadata import: ", ff)
    md <- tryCatch(read_table_flexible(ff), error = function(e) NULL)
    if (is.null(md) || nrow(md) < 10 || ncol(md) < 2) next

    candidate_cols <- names(md)[grepl("barcode|cell|cell_id|cellid|CellID|NAME", names(md), ignore.case = TRUE)]
    candidate_cols <- unique(c(candidate_cols, names(md)[1]))
    best_col <- NA_character_; best_n <- 0L
    for (cc in candidate_cols) {
      vals <- as.character(md[[cc]])
      n1 <- sum(vals %in% colnames(seu))
      n2 <- sum(paste0(basename(dirname(ff)), "_", vals) %in% colnames(seu))
      n <- max(n1, n2)
      if (n > best_n) { best_n <- n; best_col <- cc }
    }
    if (is.na(best_col) || best_n < 10) next

    cell_ids <- as.character(md[[best_col]])
    if (sum(cell_ids %in% colnames(seu)) < 10) {
      prefixed <- paste0(basename(dirname(ff)), "_", cell_ids)
      if (sum(prefixed %in% colnames(seu)) > sum(cell_ids %in% colnames(seu))) cell_ids <- prefixed
    }
    rownames(md) <- make.unique(cell_ids)
    md[[best_col]] <- NULL
    md <- md[intersect(rownames(md), colnames(seu)), , drop = FALSE]
    if (nrow(md) >= 10) {
      seu <- Seurat::AddMetaData(seu, metadata = md)
      message("Added metadata from ", ff, " for ", nrow(md), " cells.")
      break
    }
  }
  seu
}

standardize_gse176078_annotations <- function(seu) {
  md <- seu@meta.data
  if (!"celltype_subset" %in% colnames(md)) {
    exact_labels <- c("Luminal Progenitors", "Mature Luminal", "Cancer LumA SC", "Basal SC")
    exact_hit <- NA_character_
    for (cc in colnames(md)) {
      vals <- as.character(md[[cc]])
      if (any(exact_labels %in% vals, na.rm = TRUE)) { exact_hit <- cc; break }
    }
    candidates <- c(
      exact_hit, "celltype_subset", "celltype_sub", "cell_type_subset", "celltype_minor",
      "cell_type", "celltype", "CellType", "annotation", "major_celltype",
      "celltype_major", "SCSubtype", "subtype", "cell_type_major"
    )
    candidates <- candidates[!is.na(candidates)]
    ct_col <- candidates[candidates %in% colnames(md)][1]
    if (!is.na(ct_col)) {
      seu$celltype_subset <- as.character(md[[ct_col]])
      message("Standardized GSE176078 cell annotation column to celltype_subset from: ", ct_col)
    } else {
      warning("Could not identify the Wu et al. annotation column. Downstream single-cell sections may skip or fail unless celltype_subset is present.")
    }
  }

  # Make common slight naming variants compatible with the downstream manuscript tests.
  if ("celltype_subset" %in% colnames(seu@meta.data)) {
    x <- as.character(seu$celltype_subset)
    x <- dplyr::case_when(
      x %in% c("Luminal progenitors", "Luminal progenitor", "Luminal Progenitor", "LP") ~ "Luminal Progenitors",
      x %in% c("Mature luminal", "Mature Luminal Cells", "ML") ~ "Mature Luminal",
      x %in% c("Cancer LumA", "Cancer_LumA_SC", "Cancer LumA", "LumA SC", "Luminal A SC") ~ "Cancer LumA SC",
      TRUE ~ x
    )
    seu$celltype_subset <- x
  }
  seu
}

normalize_gse176078_seurat <- function(seu) {
  # Ensure a usable normalized RNA data matrix for FetchData().
  if ("RNA" %in% names(seu@assays)) Seurat::DefaultAssay(seu) <- "RNA"
  seu <- tryCatch({
    Seurat::NormalizeData(seu, verbose = FALSE)
  }, error = function(e) {
    message("NormalizeData skipped/failed; keeping loaded object as-is. Error: ", conditionMessage(e))
    seu
  })
  seu
}

build_gse176078_seurat_cache <- function() {
  sf <- params$seurat_object_path
  if (file.exists(sf) && !isTRUE(params$force_rebuild_cache)) {
    message("Using cached GSE176078 Seurat object: ", sf)
    return(invisible(TRUE))
  }
  if (!isTRUE(params$build_gse176078_seurat_if_missing)) {
    message("GSE176078 Seurat object missing, but params$build_gse176078_seurat_if_missing is FALSE; skipping.")
    return(invisible(FALSE))
  }
  if (!requireNamespace("Seurat", quietly = TRUE)) {
    if (isTRUE(params$install_missing_packages)) ensure_packages("Seurat", bioc = FALSE)
    else stop("Seurat is required to reconstruct the GSE176078 object. Install Seurat or knit with params$install_missing_packages = TRUE.", call. = FALSE)
  }

  ok <- fetch_gse176078_supplementary()
  if (!isTRUE(ok)) return(invisible(FALSE))

  gse_dir <- gse176078_dir()
  files <- decompress_gse176078_files(gse_dir)
  if (length(files) == 0) {
    stop("No GSE176078 supplementary files found under ", gse_dir, call. = FALSE)
  }

  message("Reconstructing GSE176078 Seurat object from detected processed files...")
  seu <- find_seurat_object_in_serialized_files(files)
  if (is.null(seu)) seu <- load_h5ad_as_seurat(files)
  if (is.null(seu)) seu <- load_10x_h5_as_seurat(files)
  if (is.null(seu)) seu <- load_10x_dirs_as_seurat(files)
  if (is.null(seu)) seu <- load_delimited_expression_as_seurat(files)

  if (is.null(seu)) {
    stop(
      "Could not automatically reconstruct a Seurat object from GSE176078 supplementary files.\n",
      "Supported automatic inputs are: Seurat .rds/.RData, .h5ad with SeuratDisk, 10X .h5, ",
      "10X matrix.mtx/features/barcodes directories, or delimited gene-by-cell expression matrices.\n",
      "If the public portal provides a different format, convert it once to ", sf, " and rerun the Rmd.",
      call. = FALSE
    )
  }

  seu <- add_gse176078_metadata(seu, files)
  seu <- standardize_gse176078_annotations(seu)
  seu <- normalize_gse176078_seurat(seu)

  dir.create(dirname(sf), recursive = TRUE, showWarnings = FALSE)
  saveRDS(seu, sf)
  message("Saved reconstructed GSE176078 Seurat object: ", sf)
  invisible(TRUE)
}

have_core <- all(file.exists(unname(core_rds)))
if ((!have_core || isTRUE(params$force_rebuild_cache)) && isTRUE(params$download_tcga_if_missing)) {
  build_tcga_brca_cache()
} else if (!have_core) {
  stop("Required cached files are missing: ", paste(names(core_rds)[!file.exists(unname(core_rds))], collapse = ", "),
       "\nSet params$download_tcga_if_missing = TRUE or provide the RDS files under results/rds/.", call. = FALSE)
} else {
  cat("Using cached TCGA inputs under results/rds/.\n")
}

build_xcell2_cache()
build_gse176078_seurat_cache()
```

```{r load_data}
# ---- Base data built/loaded by reproducible_input_cache ----
norm_counts       <- readRDS("results/rds/norm_counts.rds")
sample_meta_tumor <- readRDS("results/rds/sample_meta.rds")
brca.clin         <- readRDS("results/rds/brca_clin.rds")

# ---- Align norm_counts columns to sample_meta (lumA_idx indexes columns positionally) ----
if (!identical(colnames(norm_counts), sample_meta_tumor$barcode)) {
  stopifnot(setequal(colnames(norm_counts), sample_meta_tumor$barcode))
  norm_counts <- norm_counts[, sample_meta_tumor$barcode]
}

foxcut_id <- grep("ENSG00000280916", rownames(norm_counts), value = TRUE)[1]
stopifnot(!is.na(foxcut_id))

# ---- FOXC1: recover from full_counts if filtered out of norm_counts ----
foxc1_id <- grep("ENSG00000054598", rownames(norm_counts), value = TRUE)[1]
if (is.na(foxc1_id)) {
  full_counts <- readRDS("results/rds/full_counts.rds")          # adjust path if needed
  stopifnot(setequal(colnames(full_counts), colnames(norm_counts)))
  full_counts <- full_counts[, colnames(norm_counts)]            # align columns before rbind
  fc_id <- grep("ENSG00000054598", rownames(full_counts), value = TRUE)[1]
  stopifnot(!is.na(fc_id))
  norm_counts <- rbind(norm_counts, full_counts[fc_id, , drop = FALSE])
  foxc1_id <- fc_id
}
stopifnot(!is.na(foxc1_id))
```

```{r helpers}
# ---- Gene symbol lookup ----
ens_clean   <- gsub("\\..*", "", rownames(norm_counts))   # strip Ensembl version
gene_id_map <- AnnotationDbi::mapIds(
  org.Hs.eg.db, keys = ens_clean,
  column = "SYMBOL", keytype = "ENSEMBL", multiVals = "first")
names(gene_id_map) <- rownames(norm_counts)   # restore original ids for lookup_id matching
lookup_id <- function(symbol) {
  hit <- names(gene_id_map)[which(gene_id_map == symbol)]
  hit <- hit[hit %in% rownames(norm_counts)]
  if (length(hit) == 0) return(NA_character_); hit[1]
}
get_ids <- function(symbols) { ids <- sapply(symbols, lookup_id); ids[!is.na(ids)] }

# Canonical luminal progenitor panel (validated markers only; used throughout)
LP_PANEL <- c("KIT", "ALDH1A3", "PROM1", "KRT15", "SLPI", "ELF5")

make_score <- function(genes, barcodes) {
  ids <- get_ids(genes)
  scale(colMeans(log2(norm_counts[ids, barcodes] + 1), na.rm = TRUE))[, 1]
}

cn <- "groupCentaur"
extract_hr <- function(m, coef_name = cn) {
  s <- summary(m)
  if (!coef_name %in% rownames(s$conf.int))
    return(data.frame(HR = NA, Lower = NA, Upper = NA, p = NA))
  data.frame(HR = round(s$conf.int[coef_name,1],3),
             Lower = round(s$conf.int[coef_name,3],3),
             Upper = round(s$conf.int[coef_name,4],3),
             p = signif(s$coefficients[coef_name,5],3))
}
```

```{r groups}
# ---- Indices ----
lumA_idx  <- sample_meta_tumor$paper_BRCA_Subtype_PAM50 == "LumA"
basal_idx <- sample_meta_tumor$paper_BRCA_Subtype_PAM50 == "Basal"
lumB_idx  <- sample_meta_tumor$paper_BRCA_Subtype_PAM50 == "LumB"
for (v in c("lumA_idx","basal_idx","lumB_idx")) {
  x <- get(v); x[is.na(x)] <- FALSE; assign(v, x)
}

foxc1_all  <- as.numeric(norm_counts[foxc1_id, ])
foxcut_all <- as.numeric(norm_counts[foxcut_id, ])
foxc1_lumA  <- foxc1_all[lumA_idx]
foxcut_lumA <- foxcut_all[lumA_idx]

# ---- Median split (progenitor cell-of-origin boundary) ----
foxc1_med <- median(foxc1_all[lumA_idx], na.rm = TRUE)
CENTAUR_LOG2 <- round(log2(foxc1_med + 1), 2)

group_df <- data.frame(
  barcode = sample_meta_tumor$barcode, patient = sample_meta_tumor$patient,
  pam50 = sample_meta_tumor$paper_BRCA_Subtype_PAM50,
  foxc1 = foxc1_all, foxcut = foxcut_all, stringsAsFactors = FALSE)
group_df$group <- NA_character_
group_df$group[group_df$pam50=="LumA" & group_df$foxc1>=foxc1_med] <- "Centaur"
group_df$group[group_df$pam50=="LumA" & group_df$foxc1< foxc1_med] <- "LumA_non_centaur"
group_df$group[group_df$pam50=="Basal"] <- "Basal"
group_df$group[group_df$pam50=="LumB"]  <- "LumB"
group_df <- group_df[!is.na(group_df$group), ]
group_df$group <- factor(group_df$group,
  levels = c("LumA_non_centaur","Centaur","LumB","Basal"))
grp_colors <- c(LumA_non_centaur="#377eb8", Centaur="#984ea3",
                 LumB="#4daf4a", Basal="#e41a1c")
```

```{r survival_construction}
# ---- Survival data: censored (4000d) and uncensored ----
build_surv <- function(censor = NULL) {
  s <- merge(group_df[group_df$pam50=="LumA", ],
    brca.clin[, c("submitter_id","vital_status","days_to_death",
       "days_to_last_follow_up","age_at_diagnosis","ajcc_pathologic_stage",
       "treatments_radiation_treatment_or_therapy")],
    by.x="patient", by.y="submitter_id")
  s$time  <- ifelse(s$vital_status=="Dead", s$days_to_death, s$days_to_last_follow_up)
  s$event <- as.integer(s$vital_status=="Dead")
  s <- s[!is.na(s$time) & s$time > 0, ]
  if (!is.null(censor)) {
    s$event <- ifelse(s$time > censor, 0, s$event)
    s$time  <- pmin(s$time, censor)
  }
  s$age <- as.numeric(s$age_at_diagnosis)/365.25
  s$stage_clean <- gsub("[ABC]$","",gsub("Stage ","",s$ajcc_pathologic_stage))
  s$stage_clean[s$stage_clean %in% c("","X","NA")] <- NA
  s$radiation <- s$treatments_radiation_treatment_or_therapy
  s
}
surv_all   <- build_surv(censor = 4000)
surv_unc   <- build_surv(censor = NULL)
surv_norad <- surv_all[surv_all$radiation=="no" & !is.na(surv_all$radiation), ]
surv_rad   <- surv_all[surv_all$radiation=="yes" & !is.na(surv_all$radiation), ]
surv_norad_unc <- surv_unc[surv_unc$radiation=="no" & !is.na(surv_unc$radiation), ]

# Composite scores on no-rad cohort
surv_norad$prog_z  <- make_score(LP_PANEL, surv_norad$barcode)
surv_norad$strom_z <- make_score(c("PECAM1","CDH5","PDGFRA","ACTA2","ACKR1"), surv_norad$barcode)
surv_norad$tls_z   <- make_score(c("CCL19","CCL21","CXCL13"), surv_norad$barcode)
surv_norad$cyto_z  <- make_score(c("CD8A","CD8B","GZMA","GZMB","PRF1","GNLY","NKG7","IFNG"), surv_norad$barcode)

cat("Groups:\n"); print(table(group_df$group))
cat("\nNo-radiation cohort: n =", nrow(surv_norad), ", events =", sum(surv_norad$event), "\n")
```

```{r gene_panels}
basal_identity   <- c("KRT5","KRT14","KRT17","EGFR","FOXC1","MIA","PHGDH","CDH3",
                       "SFRP1","TP63","ACTA2","MYH11","CALD1","CNN1","S100A2","VIM","NGFR")
luminal_identity <- c("ESR1","PGR","FOXA1","GATA3","XBP1","AGR2","MLPH","SLC39A6",
                       "KRT8","KRT18","MUC1","CDH1")
proliferation    <- c("MKI67","CCNB1","CCNB2","CCNA2","PLK1","AURKA","AURKB","TOP2A",
                       "CDC20","BUB1","BUB1B","CENPE","CENPF","TTK","CDK1","CCND1","MCM2","PCNA","RRM2")
progenitor       <- LP_PANEL
immune_panel     <- c("CD8A","CD8B","PRF1","GZMA","GZMB","NKG7","GNLY","IFNG","IL2",
                       "CCL19","CCL21","CD79B","CD79A","MS4A1","CD19","CXCL9","CCL5","CCL4",
                       "PDCD1","CD274","CTLA4","TIGIT","HAVCR2","FOXP3","IL2RA",
                       "CXCL1","CCL2","CD68","ITGAM","CD163","MRC1","TNF",
                       "CXCL13","CR2","FCER2","LAMP3","CCR7")
```

---

# FOXC1 Heterogeneity in Luminal A

```{r s1_dist, fig.height=5, fig.width=12}
foxc1_df <- data.frame(pam50 = sample_meta_tumor$paper_BRCA_Subtype_PAM50,
  foxc1 = log2(foxc1_all+1), foxcut = log2(foxcut_all+1))
foxc1_df <- foxc1_df[foxc1_df$pam50 %in% c("LumA","LumB","Her2","Basal"), ]
foxc1_df$pam50 <- factor(foxc1_df$pam50, levels=c("LumA","LumB","Her2","Basal"))

# ---- FOXCUT--FOXC1 Spearman correlations cited in the text ----
# exact = FALSE is used because RNA-seq expression contains ties after normalization/log transform.
foxcut_foxc1_cor <- function(subtype_label) {
  d <- foxc1_df[foxc1_df$pam50 == subtype_label, c("foxc1", "foxcut")]
  d <- d[stats::complete.cases(d), ]
  ct <- suppressWarnings(stats::cor.test(d$foxc1, d$foxcut,
                                         method = "spearman", exact = FALSE))
  data.frame(
    Subtype = subtype_label,
    N = nrow(d),
    Spearman_rho = unname(ct$estimate),
    P_value = ct$p.value,
    stringsAsFactors = FALSE
  )
}

foxcut_foxc1_stats <- do.call(rbind, lapply(c("LumA", "Basal"), foxcut_foxc1_cor))
rho_fc <- foxcut_foxc1_stats$Spearman_rho[foxcut_foxc1_stats$Subtype == "LumA"]
p_fc   <- foxcut_foxc1_stats$P_value[foxcut_foxc1_stats$Subtype == "LumA"]

# ---- Distribution descriptors cited in the text (LumA vs Basal FOXC1) ----
lumA_log2  <- log2(foxc1_all[lumA_idx]  + 1)
basal_log2 <- log2(foxc1_all[basal_idx] + 1)
basal_decile1 <- quantile(basal_log2, 0.10, na.rm = TRUE)        # bottom decile of Basal
pct_lumA_above <- 100 * mean(lumA_log2 >= basal_decile1, na.rm = TRUE)
cat(sprintf("LumA  FOXC1 median log2 = %.2f\n",  median(lumA_log2,  na.rm = TRUE)))
cat(sprintf("Basal FOXC1 median log2 = %.2f\n",  median(basal_log2, na.rm = TRUE)))
cat(sprintf("Basal bottom-decile log2 = %.2f\n", basal_decile1))
cat(sprintf("%% of LumA at/above Basal bottom decile = %.1f%%\n", pct_lumA_above))

p1 <- ggplot(foxc1_df, aes(foxc1, fill=pam50)) +
  geom_density(alpha=0.4) +
  geom_vline(xintercept=CENTAUR_LOG2, linetype="dashed", color="red") +
  labs(title="FOXC1 across PAM50 subtypes (LumA: normal distribution)",
       x="FOXC1 log2(counts+1)", y="Density") + theme_bw(base_size=12)
p2 <- ggplot(foxc1_df[foxc1_df$pam50=="LumA",], aes(foxc1, foxcut)) +
  geom_point(alpha=0.3, size=1) + geom_smooth(method="lm", se=FALSE, color="#984ea3") +
  labs(title=paste0("FOXCUT-FOXC1 in LumA (ρ=", round(rho_fc,3), ")"),
       x="FOXC1 log2", y="FOXCUT log2") + theme_bw(base_size=12)
p1 | p2
```

```{r s1_foxcut_foxc1_correlation}
foxcut_foxc1_stats_print <- foxcut_foxc1_stats
foxcut_foxc1_stats_print$Spearman_rho <- round(foxcut_foxc1_stats_print$Spearman_rho, 3)
foxcut_foxc1_stats_print$P_value <- format.pval(foxcut_foxc1_stats_print$P_value,
                                                digits = 3, eps = 1e-300)
knitr::kable(
  foxcut_foxc1_stats_print,
  caption = "FOXCUT--FOXC1 Spearman correlations by PAM50 subtype used for Figure 1 and manuscript text"
)
cat(sprintf("FOXCUT--FOXC1 LumA Spearman rho = %.3f, p = %.3g, n = %d\n",
            rho_fc, p_fc,
            foxcut_foxc1_stats$N[foxcut_foxc1_stats$Subtype == "LumA"]))
```

```{r s1_groups}
table(group_df$group) %>% as.data.frame() %>% setNames(c("Group","N")) %>%
  mutate(Pct = round(100*N/sum(N),1)) %>%
  knitr::kable(caption = paste0("Group sizes (median split, log2=", CENTAUR_LOG2, ")"))
```

```{r s1_unimodality}
# Modality of the FOXC1 distribution in LumA. Supports the continuous-gradient framing:
# the median split is a convention, not a discovered boundary between two populations.
fc <- log2(foxc1_lumA + 1); fc <- fc[is.finite(fc)]

# Hartigan's dip test (H0: unimodal). p > 0.05 = fail to reject unimodality (weak, by itself).
if (requireNamespace("diptest", quietly = TRUE)) {
  dp <- diptest::dip.test(fc)
  cat(sprintf("Hartigan's dip test: D = %.4f, p = %.3f\n", dp$statistic, dp$p.value))
} else cat("diptest not installed.\n")

# BIC: 1- vs 2-component Gaussian mixture. This gives POSITIVE evidence for unimodality
# when the 1-component model wins (mclust BIC: higher = better fit).
if (requireNamespace("mclust", quietly = TRUE)) {
  bic <- mclust::mclustBIC(fc, G = 1:2, modelNames = "V")
  cat("\nGaussian mixture BIC (modelName V; higher = better):\n"); print(bic)
  b1 <- bic["1","V"]; b2 <- bic["2","V"]
  cat(sprintf("\nΔBIC (G1 − G2) = %+.2f  →  %s favored\n",
      b1 - b2, ifelse(b1 > b2, "1 component (unimodal)", "2 components (bimodal)")))
} else cat("mclust not installed.\n")
```

---

# Survival Analysis

## Kaplan-Meier

```{r s2_km, fig.height=14, fig.width=12}
# Median thresholds on the full LumA cohort (same convention as the FOXC1 split)
foxcut_med <- median(foxcut_lumA, na.rm = TRUE)

# Per-gene high/low strata, named after the gene (not "Centaur")
add_gene_splits <- function(df) {
  df$FOXCUT <- factor(ifelse(df$foxcut >= foxcut_med, "FOXCUT high", "FOXCUT low"),
                      levels = c("FOXCUT low", "FOXCUT high"))
  df$FOXC1  <- factor(ifelse(df$foxc1  >= foxc1_med,  "FOXC1 high",  "FOXC1 low"),
                      levels = c("FOXC1 low",  "FOXC1 high"))
  df
}
surv_norad <- add_gene_splits(surv_norad)
surv_rad   <- add_gene_splits(surv_rad)

# KM stratified by an explicit gene split; strata labelled by gene level
make_km_gene <- function(df, gene, title_str) {
  df$grp <- df[[gene]]
  fit <- survfit(Surv(time, event) ~ grp, data = df)
  ggsurvplot(fit, data = df, pval = TRUE, pval.method = TRUE,
    risk.table = TRUE, risk.table.height = 0.28, conf.int = TRUE, conf.int.alpha = 0.12,
    palette = c("#377eb8", "#984ea3"),
    legend.title = gene, legend.labs = levels(df$grp),
    xlab = "Days", ylab = "Overall Survival", title = title_str,
    break.time.by = 1000, ggtheme = theme_bw(base_size = 12))
}

km_foxcut_norad <- make_km_gene(surv_norad, "FOXCUT", "FOXCUT (radiation-naive)")
km_foxc1_norad  <- make_km_gene(surv_norad, "FOXC1",  "FOXC1 (radiation-naive)")
km_foxcut_rad   <- make_km_gene(surv_rad,   "FOXCUT", "FOXCUT (irradiated)")
km_foxc1_rad    <- make_km_gene(surv_rad,   "FOXC1",  "FOXC1 (irradiated)")

arrange_ggsurvplots(
  list(km_foxcut_norad, km_foxc1_norad, km_foxcut_rad, km_foxc1_rad),
  ncol = 2, nrow = 2)
```

```{r s2_cox_genes}
gene_cox <- function(df, gene, cohort) {
  df$grp <- df[[gene]]   # "<gene> low" is the reference level
  m <- coxph(Surv(time, event) ~ grp, data = df)
  # coefficient is named "grp<high level>" (e.g. "grpFOXC1 high"), not "groupCentaur"
  cbind(Gene = gene, Cohort = cohort,
        extract_hr(m, paste0("grp", levels(df$grp)[2])))
}
rbind(
  gene_cox(surv_norad, "FOXCUT", "No-radiation"),
  gene_cox(surv_norad, "FOXC1",  "No-radiation"),
  gene_cox(surv_rad,   "FOXCUT", "Radiation"),
  gene_cox(surv_rad,   "FOXC1",  "Radiation")
) %>% knitr::kable(caption = "Per-gene median-split Cox HR (high vs low) by radiation status")
```

## Cox & Censoring Sensitivity

```{r s2_cox}
rbind(
  cbind(Cohort="No-radiation", extract_hr(coxph(Surv(time,event)~group, data=surv_norad))),
  cbind(Cohort="Radiation",    extract_hr(coxph(Surv(time,event)~group, data=surv_rad)))
) %>% knitr::kable(caption="Centaur Cox HR by radiation status")
```

```{r s2_censor_sensitivity}
# Number at risk + sensitivity across censoring windows
fit_unc <- survfit(Surv(time,event)~group, data=surv_norad_unc)
cat("Number at risk by group at days 2000/3000/4000/5000/6000:\n")
print(summary(fit_unc, times=c(2000,3000,4000,5000,6000))$n.risk)

sens <- data.frame()
for (cens in c(3000,3500,4000,4500,5000,Inf)) {
  d <- surv_norad_unc
  d$e2 <- ifelse(d$time>cens, 0, d$event); d$t2 <- pmin(d$time, cens)
  m <- coxph(Surv(t2,e2)~group, data=d); s <- summary(m)
  sens <- rbind(sens, data.frame(Censor=ifelse(is.infinite(cens),"Inf",cens),
    HR=round(s$conf.int[1,1],3), p=signif(s$coefficients[1,5],3), Events=m$nevent))
}
knitr::kable(sens, caption="Censoring sensitivity (HR stable; no events past 4000d)")

cat("\nProportional hazards test:\n")
print(cox.zph(coxph(Surv(time,event)~group, data=surv_norad_unc)))
```

## Radiation × Genotype Interaction

```{r s2_interaction}
# Formal genotype × radiation interaction (paper cites interaction p < 0.01).
# Pooled LumA cohort restricted to patients with known radiation status.
surv_int <- surv_all[surv_all$radiation %in% c("yes","no"), ]
surv_int <- add_gene_splits(surv_int)
surv_int$radiation <- factor(surv_int$radiation, levels = c("no","yes"))
surv_int$centaur   <- factor(ifelse(surv_int$group == "Centaur","Centaur","Non_centaur"),
                             levels = c("Non_centaur","Centaur"))

# Likelihood-ratio test on the interaction term (additive vs full model)
int_test <- function(var) {
  m_full <- coxph(as.formula(paste0("Surv(time,event) ~ ", var, " * radiation")), data = surv_int)
  m_add  <- coxph(as.formula(paste0("Surv(time,event) ~ ", var, " + radiation")), data = surv_int)
  lrt    <- anova(m_add, m_full)
  data.frame(Predictor = var,
             Interaction_p_LRT = signif(lrt[[ncol(lrt)]][2], 3))   # last col, 2nd row = p
}
rbind(int_test("FOXC1"), int_test("FOXCUT"), int_test("centaur")) %>%
  knitr::kable(caption = "Likelihood-ratio test for genotype × radiation interaction (pooled LumA, known radiation status)")

# Full FOXC1 × radiation model coefficients (source of the in-text stratum HRs)
cat("\nFOXC1 × radiation interaction model:\n")
print(summary(coxph(Surv(time,event) ~ FOXC1 * radiation, data = surv_int))$coefficients)
```

## Continuous (Non-Dichotomized) Cox

```{r s2_continuous}
# Continuous per-SD Cox in the radiation-naive cohort -- avoids the median split and
# gives the more stable estimate referenced in the Limitations.
surv_norad$foxc1_z  <- scale(log2(surv_norad$foxc1  + 1))[,1]
surv_norad$foxcut_z <- scale(log2(surv_norad$foxcut + 1))[,1]
rbind(
  cbind(Gene = "FOXC1 (per SD, continuous)",
        extract_hr(coxph(Surv(time,event) ~ foxc1_z,  data = surv_norad), "foxc1_z")),
  cbind(Gene = "FOXCUT (per SD, continuous)",
        extract_hr(coxph(Surv(time,event) ~ foxcut_z, data = surv_norad), "foxcut_z"))
) %>% knitr::kable(caption = "Continuous per-SD Cox HR, radiation-naive (no dichotomization)")
```

## Spline Non-Linearity Check (backs the continuous-axis / not-a-threshold claim)

```{r s2_spline, eval=requireNamespace("rms", quietly=TRUE)}
# [ADDED] Confirms the FOXC1 survival effect is continuous/monotonic rather than threshold-driven:
# the non-linear spline terms should be non-significant. (Earlier cutpoint-scanning at ~Q0.45 was a
# methodological artifact; this is the principled check.)
suppressMessages(library(rms))
.sdf <- surv_norad
.sdf$log_foxc1 <- log2(.sdf$foxc1 + 1)
dd <- datadist(.sdf); options(datadist = "dd")
.fit_rcs <- cph(Surv(time, event) ~ rcs(log_foxc1, 4), data = .sdf, x = TRUE, y = TRUE)
print(anova(.fit_rcs))   # inspect the 'Nonlinear' row p (expected ns -> continuous effect)
```

---

# Immune Gene Correlations

```{r s3_immune, fig.height=8, fig.width=10}
immune_ids <- get_ids(immune_panel)
imm <- data.frame()
for (sym in names(immune_ids)) {
  ct <- cor.test(foxc1_lumA, as.numeric(norm_counts[immune_ids[sym], lumA_idx]),
                  method="spearman", exact=FALSE)
  imm <- rbind(imm, data.frame(gene=sym, rho=round(ct$estimate,3), p=signif(ct$p.value,3)))
}
imm$fdr <- p.adjust(imm$p, "BH"); imm <- imm[order(-imm$rho), ]
imm$gene <- factor(imm$gene, levels=rev(imm$gene))
imm$cat <- case_when(
  imm$gene %in% c("CCL19","CCL21","CXCL13") ~ "TLS",
  imm$gene %in% c("CD8A","CD8B","PRF1","GZMA","GZMB","NKG7","GNLY","IFNG","IL2") ~ "Cytotoxic",
  imm$gene %in% c("CD79B","CD79A","MS4A1","CD19") ~ "B cell",
  imm$gene %in% c("CR2","FCER2") ~ "FDC",
  imm$gene %in% c("PDCD1","CD274","CTLA4","TIGIT","HAVCR2") ~ "Checkpoint",
  imm$gene %in% c("FOXP3","IL2RA") ~ "Treg",
  imm$gene %in% c("CXCL1","CCL2","CD68","ITGAM","CD163","MRC1") ~ "Myeloid", TRUE ~ "Other")

# ---- Table: every immune-gene correlation the paper cites (Section 3) ----
imm_tab <- imm[order(-imm$rho), c("gene","cat","rho","p","fdr")]
imm_tab$gene <- as.character(imm_tab$gene)
imm_tab$fdr  <- signif(imm_tab$fdr, 3)
knitr::kable(imm_tab, row.names = FALSE,
  caption = "FOXC1 \u00d7 immune-gene Spearman correlations in LumA (Section 3 panel; BH-corrected)")

fig_immune <- ggplot(imm, aes(rho, gene, fill=cat)) + geom_col() + geom_vline(xintercept=0) +
  labs(title="FOXC1 × immune correlations in LumA", x="Spearman ρ", y=NULL, fill=NULL) +
  theme_bw(base_size=11)
fig_immune
```

---

# Stromal Gene Correlations

```{r s4_stromal, fig.height=9, fig.width=10}
stromal_panel <- c("ACKR1","SELP","SELE","VCAM1","ICAM1","PECAM1","CDH5","TEK","VWF","PLVAP",
  "VEGFA","PDGFRA","ACTA2","PDGFRB","MMP2","COL1A1","COL1A2","COL3A1","MCAM","RGS5",
  "PDPN","IL7","CXCL12","LUM","DCN","APOD","DES","LTB","LTA","LTBR",
  "TNFSF13B","TNFSF13","CR2","LAMP3","CCR7")
strom_ids <- get_ids(unique(stromal_panel))
strom <- data.frame()
for (sym in names(strom_ids)) {
  ct1 <- cor.test(foxc1_lumA,  as.numeric(norm_counts[strom_ids[sym], lumA_idx]), method="spearman", exact=FALSE)
  ct2 <- cor.test(foxcut_lumA, as.numeric(norm_counts[strom_ids[sym], lumA_idx]), method="spearman", exact=FALSE)
  strom <- rbind(strom, data.frame(gene=sym,
    foxc1=round(ct1$estimate,3),  foxc1_p=ct1$p.value,
    foxcut=round(ct2$estimate,3), foxcut_p=ct2$p.value))
}
# [REVIEW FIX] BH-FDR within the stromal panel (previously rho only -- no p, no correction,
# yet Section 4 of the manuscript decorates these with *** ; regenerate the stars from *_fdr).
strom$foxc1_fdr  <- p.adjust(strom$foxc1_p,  "BH")
strom$foxcut_fdr <- p.adjust(strom$foxcut_p, "BH")
strom <- strom[order(-strom$foxc1), ]

# ---- Table: every stromal-gene correlation the paper cites (Section 4) ----
knitr::kable(transform(strom,
    foxc1_p=signif(foxc1_p,3),  foxc1_fdr=signif(foxc1_fdr,3),
    foxcut_p=signif(foxcut_p,3), foxcut_fdr=signif(foxcut_fdr,3)),
  row.names = FALSE,
  caption = "FOXC1/FOXCUT \u00d7 stromal-gene Spearman correlations in LumA (Section 4; BH-corrected within panel)")

fig_stromal <- strom %>% pivot_longer(c(foxc1,foxcut), names_to="src", values_to="rho") %>%
  ggplot(aes(rho, reorder(gene, rho), fill=src)) +
  geom_col(position=position_dodge(0.7), width=0.6) + geom_vline(xintercept=0) +
  scale_fill_manual(values=c(foxc1="#2c3e50", foxcut="#e74c3c")) +
  labs(title="Stromal correlations (FOXC1 vs FOXCUT)", x="Spearman ρ", y=NULL, fill=NULL) +
  theme_bw(base_size=10)
fig_stromal
```

---

# xCell2 Deconvolution

```{r s5_xcell2, fig.height=7, fig.width=9}
xf <- "results/rds/xcell2_scores.rds"
fig_xcell2 <- NULL
if (!file.exists(xf)) {
  cat("xCell2 scores not found at", xf, "- skipping.\n")
} else {
  xc <- readRDS(xf)
  samp <- sample_meta_tumor$barcode[lumA_idx]
  # Orient to cell-types (rows) x samples (cols): TCGA barcodes mark the sample dim
  bc_in_cols <- any(grepl("^TCGA", colnames(xc)))
  bc_in_rows <- any(grepl("^TCGA", rownames(xc)))
  if (bc_in_rows && !bc_in_cols) xc <- t(xc)   # barcodes were in rows -> move to columns

  # Harmonize barcode formats (dots->dashes, find matching TCGA level)
  norm_bc <- function(x) toupper(gsub("\\.", "-", x))
  samp_n <- norm_bc(samp); xc_n <- norm_bc(colnames(xc))
  klen <- NA
  for (k in c(28, 19, 16, 15, 12)) {
    if (length(intersect(substr(samp_n, 1, k), substr(xc_n, 1, k))) > 0) { klen <- k; break }
  }

  if (is.na(klen)) {
    cat("xCell2 barcodes do not match sample_meta - skipping.\n")
    cat("  xc cols:", paste(head(colnames(xc), 4), collapse = " | "), "\n")
    cat("  sample :", paste(head(samp, 4), collapse = " | "), "\n")
  } else {
    samp_k <- substr(samp_n, 1, klen); xc_k <- substr(xc_n, 1, klen)
    fdf <- data.frame(id = samp_k, foxc1 = foxc1_lumA, stringsAsFactors = FALSE)
    fdf <- fdf[!duplicated(fdf$id), ]
    keep <- !duplicated(xc_k); xc <- xc[, keep, drop = FALSE]; xc_k <- xc_k[keep]
    common <- intersect(fdf$id, xc_k)
    fx  <- fdf$foxc1[match(common, fdf$id)]
    xcm <- xc[, match(common, xc_k), drop = FALSE]

    xres <- data.frame()
    for (ct in rownames(xcm)) {
      m <- as.numeric(xcm[ct, ])
      if (all(is.na(m))) next
      r <- suppressWarnings(cor.test(fx, m, method = "spearman", exact = FALSE))
      xres <- rbind(xres, data.frame(cell_type = ct, rho = round(r$estimate, 3),
                                     p = signif(r$p.value, 3)))
    }
    xres$fdr <- signif(p.adjust(xres$p, "BH"), 3)
    xres <- xres[order(-xres$rho), ]

    # ---- Bar graph: top cell types by |rho|, colored by direction ----
    top <- xres[order(-abs(xres$rho)), ][seq_len(min(20, nrow(xres))), ]
    fig_xcell2 <- ggplot(top, aes(rho, reorder(cell_type, rho), fill = rho > 0)) +
        geom_col(width = 0.7) + geom_vline(xintercept = 0) +
        scale_fill_manual(values = c(`TRUE` = "#984ea3", `FALSE` = "#377eb8"),
                          labels = c(`TRUE` = "Positive (centaur-enriched)",
                                     `FALSE` = "Negative"),
                          name = NULL) +
        labs(title = "FOXC1 correlation with xCell2 cell-type enrichment (LumA)",
             subtitle = paste0("Top ", nrow(top),
                               " cell types by |\u03c1|; matched at barcode length ", klen),
             x = "Spearman \u03c1 with FOXC1", y = NULL) +
        theme_bw(base_size = 11) + theme(legend.position = "bottom")
    print(fig_xcell2)
    knitr::kable(head(xres, 20),
                 caption = "Top FOXC1-correlated cell types (xCell2, BH-corrected)")
  }
}
```

---

# The Molecular Centaur

## Dual Lineage: Basal + Luminal

```{r s6_lineage, fig.height=8, fig.width=12}
basal_ids <- get_ids(basal_identity); lum_ids <- get_ids(luminal_identity)
all_ids <- c(basal_ids, lum_ids); all_nm <- c(names(basal_ids), names(lum_ids))
ez <- t(scale(t(log2(norm_counts[all_ids, group_df$barcode]+1))))
rownames(ez) <- all_nm; ez[ez>2] <- 2; ez[ez< -2] <- -2
ca <- data.frame(Group=group_df$group, row.names=group_df$barcode)
ra <- data.frame(Program=c(rep("Basal",length(basal_ids)), rep("Luminal",length(lum_ids))),
  row.names=all_nm)
ph_lineage <- pheatmap(ez[, order(ca$Group)], annotation_col=ca, annotation_row=ra,
  cluster_cols=FALSE, show_colnames=FALSE, fontsize_row=7,
  color=colorRampPalette(c("navy","white","firebrick"))(100),
  annotation_colors=list(Group=grp_colors, Program=c(Basal="#e41a1c", Luminal="#377eb8")),
  main="Dual lineage: basal elevated, luminal retained")
```

### Lineage-marker differential expression with FDR (backs "16 of 17" and "11 of 12")

```{r s6_lineage_differential}
# [ADDED] The manuscript cites "16 of 17 basal markers FDR<0.05 (Centaur vs non-Centaur)" and
# "11 of 12 luminal markers higher than Basal (p<0.05)", but these per-marker tests were not in
# the prior pipeline. Computed here with BH-FDR within each family.
.bc <- group_df$barcode; .grp <- group_df$group
.lin_one <- function(sym) {
  gid <- get_ids(sym); if (length(gid) == 0) return(NULL)
  x   <- log2(as.numeric(norm_counts[gid[1], .bc]) + 1)
  cen <- x[.grp=="Centaur"]; non <- x[.grp=="LumA_non_centaur"]; bas <- x[.grp=="Basal"]
  data.frame(gene=sym,
    med_cen=round(median(cen),2), med_non=round(median(non),2), med_bas=round(median(bas),2),
    p_cen_vs_non = wilcox.test(cen, non)$p.value,
    p_cen_vs_bas = wilcox.test(cen, bas)$p.value)
}
basal_de <- do.call(rbind, lapply(basal_identity,   .lin_one))
lum_de   <- do.call(rbind, lapply(luminal_identity, .lin_one))
basal_de$fdr_cen_vs_non <- p.adjust(basal_de$p_cen_vs_non, "BH")
lum_de$fdr_cen_vs_bas   <- p.adjust(lum_de$p_cen_vs_bas,   "BH")

# NOTE: basal_identity INCLUDES FOXC1, and Centaur is defined by FOXC1-high -> FOXC1's significance
# here is circular. Exclude it from the tally you report (so it is "out of 16", not 17).
n_basal_up <- sum(basal_de$gene != "FOXC1" & basal_de$med_cen > basal_de$med_non &
                  basal_de$fdr_cen_vs_non < 0.05)
cat(sprintf("Basal markers (excl. FOXC1) elevated in Centaur vs non-Centaur at FDR<0.05: %d / %d\n",
    n_basal_up, sum(basal_de$gene != "FOXC1")))
n_lum_ret <- sum(lum_de$med_cen > lum_de$med_bas & lum_de$p_cen_vs_bas < 0.05)
cat(sprintf("Luminal markers retained (Centaur > Basal) at p<0.05: %d / %d\n",
    n_lum_ret, nrow(lum_de)))
knitr::kable(transform(basal_de,
    p_cen_vs_non=signif(p_cen_vs_non,3), fdr_cen_vs_non=signif(fdr_cen_vs_non,3),
    p_cen_vs_bas=signif(p_cen_vs_bas,3)),
  row.names=FALSE, caption="Basal-identity markers: Centaur vs non-Centaur LumA (BH within panel)")
knitr::kable(transform(lum_de,
    p_cen_vs_non=signif(p_cen_vs_non,3),
    p_cen_vs_bas=signif(p_cen_vs_bas,3), fdr_cen_vs_bas=signif(fdr_cen_vs_bas,3)),
  row.names=FALSE, caption="Luminal-identity markers: Centaur vs Basal (BH within panel)")
```

```{r s6_ssgsea, fig.height=6, fig.width=12, eval=has_gsva}
library(GSVA)
sf <- "results/rds/hybrid_ssgsea.rds"
if (!file.exists(sf)) {
  sigs <- list(basal=basal_ids, luminal=lum_ids)
  par <- ssgseaParam(exprData=as.matrix(log2(norm_counts+1)), geneSets=sigs)
  ss <- gsva(par, verbose=FALSE); saveRDS(ss, sf)
} else ss <- readRDS(sf)
sdf <- data.frame(barcode=colnames(ss), basal=ss["basal",], luminal=ss["luminal",])
sdf <- merge(sdf, group_df[,c("barcode","group")], by="barcode")
fig_ssgsea <- ggplot(sdf, aes(luminal, basal, color=group)) + geom_point(alpha=0.5, size=1.5) +
  stat_ellipse(level=0.68, linewidth=1) + scale_color_manual(values=grp_colors) +
  labs(title="2D lineage identity space (centaur: basal up, luminal flat)",
       x="Luminal ssGSEA", y="Basal ssGSEA") + theme_bw(base_size=12)
fig_ssgsea
```

## Proliferative Quiescence (Luminal-Associated)

```{r s6_prolif, fig.height=6, fig.width=10}
prolif_ids <- get_ids(proliferation)
pscore <- colMeans(log2(norm_counts[prolif_ids, group_df$barcode]+1), na.rm=TRUE)
pdf <- merge(data.frame(barcode=names(pscore), prolif=pscore),
  group_df[,c("barcode","group")], by="barcode")
pcor <- data.frame()
for (sym in names(prolif_ids)) {
  ct <- cor.test(foxc1_lumA, as.numeric(norm_counts[prolif_ids[sym], lumA_idx]),
                  method="spearman", exact=FALSE)
  pcor <- rbind(pcor, data.frame(gene=sym, rho=round(ct$estimate,3)))
}
fig_prolif_box <- ggplot(pdf, aes(group, prolif, fill=group)) +
  geom_violin(alpha=0.6) + geom_boxplot(width=0.15, outlier.size=0.5) +
  scale_fill_manual(values=grp_colors) +
  labs(title="Proliferation by group", subtitle="Centaur ≈ non-centaur << LumB < Basal",
       x=NULL, y="Mean log2 prolif") + theme_bw(base_size=12) + theme(legend.position="none")
fig_prolif_cor <- ggplot(pcor, aes(rho, reorder(gene, rho))) + geom_col(fill="#7f8c8d") +
  geom_vline(xintercept=c(-0.3,0,0.3), linetype=c("dashed","solid","dashed")) +
  labs(title=paste("Mean ρ =", round(mean(pcor$rho),3)), x="ρ with FOXC1", y=NULL) +
  theme_bw(base_size=10)
fig_prolif_box | fig_prolif_cor
```

## Progenitor Markers

```{r s6_prog, fig.height=4, fig.width=9}
# LP_PANEL (6 genes) is defined globally and is the COMPOSITE used in mediation (prog_z).
# Do NOT add genes to LP_PANEL here -- doing so would change the mediation HR in s7.
# For reporting we ALSO correlate the negative markers CD24/EPCAM/CD44 -- correlation
# only, NOT folded into the composite.
prog_pos <- LP_PANEL                      # positive progenitor markers (validated panel)
prog_neg <- c("CD24","EPCAM","CD44")      # negative markers, correlation only

cor_one <- function(sym) {
  gid <- lookup_id(sym); if (is.na(gid)) return(NULL)
  ct <- cor.test(foxc1_lumA, as.numeric(norm_counts[gid, lumA_idx]),
                  method="spearman", exact=FALSE)
  data.frame(gene=sym, rho=round(ct$estimate,3), p=signif(ct$p.value,3))
}
prog <- do.call(rbind, lapply(c(prog_pos, prog_neg), cor_one))
prog$direction <- ifelse(prog$gene %in% prog_neg, "negative", "positive")
prog$fdr <- signif(p.adjust(prog$p, "BH"), 3)
prog <- prog[order(-prog$rho), ]

# Output the table
knitr::kable(prog, caption="FOXC1 × progenitor markers in LumA (BH-corrected; negative markers are reported but NOT in the mediation composite)")

# Then the plot
fig_prog <- ggplot(prog, aes(rho, reorder(gene, rho), fill=direction)) +
  geom_col() + geom_vline(xintercept=0) +
  scale_fill_manual(values=c(positive="#2c3e50", negative="#c0392b")) +
  labs(title="Progenitor markers vs FOXC1 in LumA", x="Spearman ρ", y=NULL, fill=NULL) +
  theme_bw(base_size=12)
fig_prog
```

## Paracrine Factors by Cellular Source

```{r s6_paracrine, fig.height=6, fig.width=10}
myo_para  <- c("FGF2","HGF","NRG1")
strom_para<- c("BMP2","BMP4","FGF7","PDGFA","PDGFB")
mixed_para<- c("IGF1","TGFB1","TGFB2","TGFB3","WNT5A")
para_all <- c(myo_para, strom_para, mixed_para)
para_ids <- get_ids(para_all)
para <- data.frame()
for (sym in names(para_ids)) {
  ct <- cor.test(foxc1_lumA, as.numeric(norm_counts[para_ids[sym], lumA_idx]),
                  method="spearman", exact=FALSE)
  vals <- log2(as.numeric(norm_counts[para_ids[sym], group_df$barcode])+1)
  meds <- tapply(vals, group_df$group, median, na.rm=TRUE)
  src <- case_when(sym %in% myo_para~"Myoepithelial", sym %in% strom_para~"Stromal",
                    TRUE~"Mixed/uncertain")
  para <- rbind(para, data.frame(gene=sym, source=src, rho=round(ct$estimate,3),
    med_centaur=round(meds["Centaur"],2), med_basal=round(meds["Basal"],2)))
}
para <- para[order(-para$rho), ]
fig_para <- ggplot(para, aes(rho, reorder(gene, rho), fill=source)) + geom_col() + geom_vline(xintercept=0) +
  scale_fill_manual(values=c(Myoepithelial="#e74c3c", Stromal="#3498db",
    `Mixed/uncertain`="#95a5a6")) +
  labs(title="Paracrine factors by normal cellular source",
       subtitle="Myoepithelial (FGF2/NRG1/HGF) exceed Basal in centaur = cancer-intrinsic",
       x="Spearman ρ with FOXC1", y=NULL, fill="Normal source") + theme_bw(base_size=11)
fig_para
knitr::kable(para, caption="Paracrine: correlation + median expression vs Basal")
```

## Three-Tier Inflammatory Regulation

```{r s6_inflam, fig.height=5, fig.width=10}
inflam <- c("IL6","CXCL8","CXCL1","CXCL2","CXCL3","CCL2","CCL7","PTGS2","IL1A","IL1B")
inf_ids <- get_ids(inflam)
inf <- data.frame()
for (sym in names(inf_ids)) {
  vals <- log2(as.numeric(norm_counts[inf_ids[sym], group_df$barcode])+1)
  meds <- tapply(vals, group_df$group, median, na.rm=TRUE)
  inf <- rbind(inf, data.frame(gene=sym,
    delta_vs_basal=round(meds["Centaur"]-meds["Basal"],2)))
}
inf$tier <- ifelse(inf$delta_vs_basal > -0.3, "Tier1: Autonomous",
  ifelse(inf$delta_vs_basal > -1.0, "Tier2: FOXC1-restrained", "Tier3: TAM-dependent"))
inf <- inf[order(inf$delta_vs_basal), ]
fig_inflam <- ggplot(inf, aes(delta_vs_basal, reorder(gene, delta_vs_basal), fill=tier)) +
  geom_col() + geom_vline(xintercept=0, linetype="dashed") +
  scale_fill_manual(values=c(`Tier1: Autonomous`="#27ae60",
    `Tier2: FOXC1-restrained`="#f39c12", `Tier3: TAM-dependent`="#e74c3c")) +
  labs(title="Inflammatory genes: Centaur vs Basal", x="Δ median log2", y=NULL, fill=NULL) +
  theme_bw(base_size=11)
fig_inflam
```

---

# Mediation & Independence

## Clinical Sequential Cox

```{r s7_clinical}
surv_cc <- surv_norad[complete.cases(surv_norad[,c("group","age","stage_clean")]), ]
m1 <- coxph(Surv(time,event)~group, data=surv_norad)
m2 <- coxph(Surv(time,event)~group+age, data=surv_cc)
m3 <- coxph(Surv(time,event)~group+stage_clean, data=surv_cc)
m4 <- coxph(Surv(time,event)~group+age+stage_clean, data=surv_cc)
m5 <- coxph(Surv(time,event)~group+age+stage_clean+cyto_z, data=surv_cc)
rbind(
  cbind(Model="Centaur only (full)", extract_hr(m1)),
  cbind(Model="+ age", extract_hr(m2)), cbind(Model="+ stage", extract_hr(m3)),
  cbind(Model="+ age + stage", extract_hr(m4)),
  cbind(Model="+ age + stage + cytotoxic", extract_hr(m5))
) %>% knitr::kable(caption="Clinical covariate adjustment")
```

## Causal Chain (Individual Mediators, age+stage controlled)

```{r s7_chain}
surv_cc$prog_z  <- make_score(LP_PANEL, surv_cc$barcode)
surv_cc$strom_z <- make_score(c("PECAM1","CDH5","PDGFRA","ACTA2","ACKR1"), surv_cc$barcode)
surv_cc$tls_z   <- make_score(c("CCL19","CCL21","CXCL13"), surv_cc$barcode)
surv_cc$cyto_z  <- make_score(c("CD8A","CD8B","GZMA","GZMB","PRF1","GNLY","NKG7","IFNG"), surv_cc$barcode)
rbind(
  cbind(Model="age+stage baseline", extract_hr(coxph(Surv(time,event)~group+age+stage_clean, data=surv_cc))),
  cbind(Model="+ progenitor", extract_hr(coxph(Surv(time,event)~group+age+stage_clean+prog_z, data=surv_cc))),
  cbind(Model="+ stromal", extract_hr(coxph(Surv(time,event)~group+age+stage_clean+strom_z, data=surv_cc))),
  cbind(Model="+ TLS", extract_hr(coxph(Surv(time,event)~group+age+stage_clean+tls_z, data=surv_cc))),
  cbind(Model="+ cytotoxic", extract_hr(coxph(Surv(time,event)~group+age+stage_clean+cyto_z, data=surv_cc)))
) %>% knitr::kable(caption="Individual mediator adjustment (each vs age+stage baseline)")
```


## Age corelations with progenitors, foxc1

```{r}
# Direct correlation of FOXC1 and progenitor markers WITH age
age_genes <- c("FOXC1", "KIT", "ALDH1A3", "PROM1", "KRT15", "SLPI", "ELF5")
age_lumA <- as.numeric(sample_meta_tumor$age_at_diagnosis[lumA_idx])/365.25
stage_num <- as.numeric(factor(gsub("[ABC]$","",gsub("Stage ","",
  sample_meta_tumor$ajcc_pathologic_stage[lumA_idx])), levels=c("I","II","III","IV")))

vb <- !is.na(age_lumA) & !is.na(stage_num)

age_cor <- data.frame()
for (sym in age_genes) {
  gid <- lookup_id(sym)
  if (is.na(gid)) next
  expr <- as.numeric(norm_counts[gid, lumA_idx])
  ct <- cor.test(expr[vb], age_lumA[vb], method = "spearman", exact = FALSE)
  age_cor <- rbind(age_cor, data.frame(
    gene = sym, rho_vs_age = round(ct$estimate, 3),
    p = signif(ct$p.value, 3)))
}
age_cor$fdr <- signif(p.adjust(age_cor$p, "BH"), 3)
knitr::kable(age_cor,
  caption = "FOXC1 and progenitor markers vs patient age (inverse = younger patients enriched)")

# Optional: visualize FOXC1 vs age directly
fig_foxc1_age <- ggplot(data.frame(foxc1 = foxc1_lumA[vb], age = age_lumA[vb]),
       aes(age, log2(foxc1 + 1))) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", color = "#984ea3") +
  labs(title = "FOXC1 declines with patient age in LumA",
       subtitle = paste0("Spearman ρ = ",
         round(cor(foxc1_lumA[vb], age_lumA[vb], method = "spearman"), 3)),
       x = "Age at diagnosis (years)", y = "FOXC1 log2") +
  theme_bw(base_size = 12)
fig_foxc1_age
```



## Age Confound & Partial Correlations

```{r s7_age, fig.height=7, fig.width=9}
age_lumA <- as.numeric(sample_meta_tumor$age_at_diagnosis[lumA_idx])/365.25
stage_num <- as.numeric(factor(gsub("[ABC]$","",gsub("Stage ","",
  sample_meta_tumor$ajcc_pathologic_stage[lumA_idx])), levels=c("I","II","III","IV")))
vb <- !is.na(age_lumA) & !is.na(stage_num)
psp <- function(x,y,c1,c2){ ok<-complete.cases(x,y,c1,c2)
  cor.test(residuals(lm(rank(x[ok])~rank(c1[ok])+rank(c2[ok]))),
           residuals(lm(rank(y[ok])~rank(c1[ok])+rank(c2[ok])))) }
kg <- c("CD8A","PRF1","CCL19","CCL21","CD79B","PDCD1",
        "ACKR1","PECAM1","PDGFRA","CXCL12",
        "FGF2","NRG1","HGF","BMP2","IGF1",
        "KIT","ALDH1A3","PROM1",
        "CD68","CXCL1")
adj <- data.frame()
for (sym in kg) { gid<-lookup_id(sym); if(is.na(gid)) next
  tg <- as.numeric(norm_counts[gid, lumA_idx])
  raw <- cor.test(foxc1_lumA[vb], tg[vb], method="spearman", exact=FALSE)
  pc <- psp(foxc1_lumA, tg, age_lumA, stage_num)
  adj <- rbind(adj, data.frame(gene=sym, raw=round(raw$estimate,3),
    adj=round(pc$estimate,3), p=signif(pc$p.value,3))) }
adj$sig <- ifelse(p.adjust(adj$p,"bonferroni")<0.05,"Sig","NS")
fig_age_partial <- ggplot(adj, aes(raw, adj, color=sig, label=gene)) + geom_point(size=3) +
  geom_text(hjust=-0.15, size=3, show.legend=FALSE) +
  geom_abline(slope=1, intercept=0, linetype="dashed", color="grey50") +
  scale_color_manual(values=c(Sig="#2c3e50", NS="#e74c3c")) +
  labs(title="Age+stage partial correlations (above diagonal = suppressor)",
       x="Raw ρ", y="Adjusted ρ") + theme_bw(base_size=12)
fig_age_partial
```

## Age-Stratified Survival

```{r s7_strat}
surv_norad$strat <- cut(surv_norad$age, c(0,50,65,100), c("<50","50-65",">65"))
as_res <- data.frame()
for (s in c("<50","50-65",">65")) {
  sub <- surv_norad[surv_norad$strat==s & !is.na(surv_norad$strat), ]
  if (sum(sub$event)>=3 && length(unique(sub$group))==2) {
    sc <- summary(coxph(Surv(time,event)~group, data=sub))
    as_res <- rbind(as_res, data.frame(Age=s, N=nrow(sub), Events=sum(sub$event),
      HR=round(sc$conf.int[1,1],2), p=signif(sc$coefficients[1,5],3)))
  }
}
knitr::kable(as_res, caption="Age-stratified survival (HR stable across strata)")
```

---

# Orthogonality to Existing Classifications

```{r s8_ortho}
foxc1_grp <- ifelse(foxc1_lumA >= foxc1_med, "Centaur", "Non-centaur")
score_of <- function(genes) colMeans(log2(norm_counts[get_ids(genes), lumA_idx]+1), na.rm=TRUE)

# Clinical tools (proliferation-dominated)
odx   <- scale(score_of(c("MKI67","AURKA","BIRC5","CCNB1","MYBL2")))[,1] -
         0.5*scale(score_of(c("ESR1","PGR","BCL2","SCUBE2")))[,1]
mamma <- score_of(c("CCNE2","MCM6","CENPA","MMP9","CXCL12","FLT1","IGFBP5","SCUBE2"))
ror   <- score_of(c("BIRC5","CCNB1","CDC20","MKI67","RRM2","TYMS","UBE2C","MYBL2"))
aurka <- as.numeric(norm_counts[lookup_id("AURKA"), lumA_idx])
# Poudel signatures
poudel_stem <- score_of(c("LGR5","ASCL2","AXIN2","SOX9","CD44","PROM1"))
poudel_inf  <- score_of(c("CD2","CD3D","CD8A","GZMB","CXCL9","CXCL10"))
poudel_mes  <- score_of(c("ZEB1","ZEB2","VIM","FN1","SPARC","COL1A1"))
# [REVIEW FIX] Lineage contrast (luminal - basal). This is a LINEAGE-IDENTITY axis, NOT tumor
# purity: it contains FOXC1, so its correlation with the Centaur axis is circular ("by definition").
# It must NOT be used as a purity/admixture control (it was previously mislabeled "purity").
lineage_contrast <- score_of(c("ESR1","PGR","FOXA1","GATA3")) -
                    score_of(c("KRT5","KRT14","KRT17","EGFR","FOXC1"))

# REAL tumor purity for the admixture confound check -- a DNA/consensus estimate INDEPENDENT of the
# lineage RNA. Priority: (1) cached rds; (2) Aran 2015 consensus purity bundled in TCGAbiolinks
# (CPE = consensus of ESTIMATE/ABSOLUTE/LUMP/IHC; ABSOLUTE = DNA copy-number), filling CPE gaps with
# ABSOLUTE; (3) NA -> dependent checks skip. The whole thing is tryCatch-wrapped so a missing package
# or a failed barcode match can NEVER halt the knit. To use a DNA-only estimate, swap CPE for ABSOLUTE.
purity_file <- "results/rds/tcga_brca_purity_cpe.rds"
.lumA_bc <- sample_meta_tumor$barcode[lumA_idx]
purity <- tryCatch({
  if (file.exists(purity_file)) {
    pv <- readRDS(purity_file)
    as.numeric(pv[ match(substr(.lumA_bc, 1, 15), substr(names(pv), 1, 15)) ])
  } else if (requireNamespace("TCGAbiolinks", quietly = TRUE)) {
    .e <- new.env(); utils::data("Tumor.purity", package = "TCGAbiolinks", envir = .e)
    tp <- .e$Tumor.purity
    num  <- function(x) suppressWarnings(as.numeric(gsub(",", ".", as.character(x))))
    cpe  <- if ("CPE" %in% names(tp))      num(tp$CPE)      else rep(NA_real_, nrow(tp))
    abs_ <- if ("ABSOLUTE" %in% names(tp)) num(tp$ABSOLUTE) else rep(NA_real_, nrow(tp))
    cpe[is.na(cpe)] <- abs_[is.na(cpe)]                       # fill CPE gaps with ABSOLUTE
    idcol <- if ("Sample.ID" %in% names(tp)) "Sample.ID" else names(tp)[1]
    pv <- stats::setNames(cpe, as.character(tp[[idcol]]))
    # Match LumA barcodes at the substring length that resolves the most samples.
    best_k <- 15; best_n <- -1
    for (k in c(16, 15, 12)) {
      n <- length(intersect(substr(.lumA_bc, 1, k), substr(names(pv), 1, k)))
      if (n > best_n) { best_n <- n; best_k <- k }
    }
    out <- as.numeric(pv[ match(substr(.lumA_bc, 1, best_k), substr(names(pv), 1, best_k)) ])
    message(sprintf("Purity: TCGAbiolinks consensus (CPE/ABSOLUTE) matched %d/%d LumA at %d-char.",
                    sum(!is.na(out)), length(.lumA_bc), best_k))
    try({ dir.create(dirname(purity_file), recursive = TRUE, showWarnings = FALSE)
          saveRDS(stats::setNames(out, .lumA_bc), purity_file) }, silent = TRUE)  # cache (best-effort)
    out
  } else {
    message("Purity: TCGAbiolinks not installed and no cached file -- admixture checks will skip.")
    rep(NA_real_, sum(lumA_idx))
  }
}, error = function(e) {
  message("Purity load failed (", conditionMessage(e), ") -- admixture checks will skip.")
  rep(NA_real_, sum(lumA_idx))
})
cat(sprintf("Real purity available for %d / %d LumA tumors.\n", sum(!is.na(purity)), length(purity)))

ortho <- data.frame(
  System=c("Oncotype DX","MammaPrint","PAM50 ROR","SCMGENE AURKA",
           "Poudel stem","Poudel inflammatory","Poudel mesenchymal","Lineage contrast (lum-basal)"),
  rho=c(
    cor(foxc1_lumA, odx, method="spearman"),
    cor(foxc1_lumA, mamma, method="spearman"),
    cor(foxc1_lumA, ror, method="spearman"),
    cor(foxc1_lumA, aurka, method="spearman"),
    cor(foxc1_lumA, poudel_stem, method="spearman"),
    cor(foxc1_lumA, poudel_inf, method="spearman"),
    cor(foxc1_lumA, poudel_mes, method="spearman"),
    cor(foxc1_lumA, lineage_contrast, method="spearman")))
ortho$rho <- round(ortho$rho, 3)
ortho$interpretation <- c(rep("Clinical tool (expect null)",4),
  rep("Poudel (expect overlap)",3), "By definition")
knitr::kable(ortho, caption="FOXC1 orthogonality to LumA classification systems")
```

```{r s8_lineage_anchors}
# Single-gene lineage anchors cited in the text (SCMGENE genes + ER axis).
# Reproduces ESR1 rho (~ -0.43) and confirms ERBB2 is uninformative in LumA.
anchor <- data.frame()
for (sym in c("ESR1","PGR","ERBB2")) {
  gid <- lookup_id(sym); if (is.na(gid)) next
  ct <- cor.test(foxc1_lumA, as.numeric(norm_counts[gid, lumA_idx]),
                 method = "spearman", exact = FALSE)
  anchor <- rbind(anchor, data.frame(gene = sym,
    rho = round(ct$estimate, 3), p = signif(ct$p.value, 3)))
}
knitr::kable(anchor, row.names = FALSE,
  caption = "FOXC1 vs single-gene lineage anchors in LumA (ESR1 is the cited -0.43)")
```

```{r s8_plot, fig.height=5, fig.width=9}
fig_ortho <- ggplot(ortho, aes(reorder(System, abs(rho)), abs(rho), fill=abs(rho)>0.3)) +
  geom_col() + coord_flip() +
  geom_hline(yintercept=0.3, linetype="dashed", color="red") +
  scale_fill_manual(values=c(`TRUE`="#e74c3c", `FALSE`="#2c3e50"), guide="none") +
  labs(title="Orthogonality: clinical tools blind to centaur",
       subtitle="Red line = substantial overlap (0.3)", x=NULL, y="|Spearman ρ|") +
  theme_bw(base_size=12)
fig_ortho
```

```{r s8_purity_survival}
# [REVIEW FIX] The (luminal - basal) score below is a LINEAGE-IDENTITY contrast, NOT tumor purity
# (it contains FOXC1). Its association with the Centaur axis is circular, so it CANNOT serve as an
# admixture control. We (a) keep it only as a labeled lineage axis, and (b) run the real admixture
# check against a TRUE purity estimate ('purity', loaded in setup) if one was supplied.
lineage_contrast_v <- (score_of(c("ESR1","PGR","FOXA1","GATA3")) -
                       score_of(c("KRT5","KRT14","KRT17","EGFR","FOXC1")))
names(lineage_contrast_v) <- sample_meta_tumor$barcode[lumA_idx]
surv_pur <- surv_norad
surv_pur$lin_z <- scale(lineage_contrast_v[surv_norad$barcode])[,1]
spL <- surv_pur[!is.na(surv_pur$lin_z), ]
cat("[lineage contrast, NOT purity] alone HR:",
    round(summary(coxph(Surv(time,event)~lin_z, data=spL))$conf.int[1,1],3),
    "(circular with the Centaur definition; reported only as a lineage axis)\n")

# --- REAL purity sensitivity (skips if no DNA/CPE purity supplied) ---
if (exists("purity") && any(!is.na(purity))) {
  pr <- purity; names(pr) <- sample_meta_tumor$barcode[lumA_idx]
  surv_pur$pur_z <- scale(pr[surv_norad$barcode])[,1]
  spR <- surv_pur[!is.na(surv_pur$pur_z), ]
  cat("[REAL purity] alone HR:", round(summary(coxph(Surv(time,event)~pur_z, data=spR))$conf.int[1,1],3), "\n")
  cat("[REAL purity] Centaur | purity HR:",
      round(summary(coxph(Surv(time,event)~pur_z+group, data=spR))$conf.int[cn,1],3),
      "(does the Centaur benefit survive a TRUE purity adjustment?)\n")
} else {
  cat("[REAL purity check skipped] supply results/rds/tcga_brca_purity_cpe.rds (ABSOLUTE/CPE) to enable.\n")
}
```
#comparison between poudel subtypes and centaur phenotype 
```{r}
# ── Load Poudel's actual per-sample heterocellular calls ──
poudel <- read.csv("poudel_heterocellular_calls.csv", stringsAsFactors = FALSE)

# Poudel uses dot-separated 12-char patient IDs (TCGA.A2.A0CO).
# Harmonize to dash format to match your barcodes.
poudel$patient12 <- gsub("\\.", "-", poudel$Sample)

# Your LumA tumors → 12-char patient ID
group_df$patient12 <- substr(group_df$patient, 1, 12)

# ── Join (LumA only) ──
merged <- merge(
  group_df[group_df$pam50 == "LumA", c("patient12", "group")],
  poudel[, c("patient12", "Heterocellular_Subtype")],
  by = "patient12")

cat("Matched LumA tumors:", nrow(merged), "\n")
cat("Subtypes present:\n"); print(table(merged$Heterocellular_Subtype))

# ── Restrict to high-confidence calls (drop mixed/low) ──
hc <- c("Enterocyte","Goblet.like","Inflammatory","Stem.like","TA")
mh <- merged[merged$Heterocellular_Subtype %in% hc, ]
cat("\nHigh-confidence LumA tumors:", nrow(mh), "\n")

# ── The decisive cross-tab ──
tab <- table(mh$Heterocellular_Subtype,
             ifelse(mh$group == "Centaur", "Centaur", "Non_centaur"))
cat("\nCentaur vs Poudel subtype:\n"); print(tab)
cat("\n% centaur within each subtype (baseline ~50%):\n")
print(round(prop.table(tab, 1), 2))

# ── The numbers that settle it ──
cat("\n% of centaur tumors that are Poudel Stem.like:",
    round(100*mean(mh$Heterocellular_Subtype[mh$group=="Centaur"]=="Stem.like"),1),"%\n")
cat("% of Poudel Stem.like that are centaur:",
    round(100*mean(mh$group[mh$Heterocellular_Subtype=="Stem.like"]=="Centaur"),1),"%\n")
chi <- chisq.test(table(mh$Heterocellular_Subtype, mh$group))
g2 <- ifelse(mh$group == "Centaur", "Centaur", "Non")
tab2 <- table(mh$Heterocellular_Subtype, g2)
chi <- chisq.test(tab2)
V <- sqrt(as.numeric(chi$statistic) / (nrow(mh) * (min(dim(tab2)) - 1)))
cat("Cramer's V:", round(V, 3), "| chi-sq p =", signif(chi$p.value, 3), "\n")

```
#orthogonality to clinical genomic risk tools
```{r}
# === s8_clinical_tools: FOXC1 vs clinical genomic risk scores ===
library(genefu)
library(org.Hs.eg.db)

# genefu expects samples x genes; build expression + annotation
lumA_expr <- t(log2(norm_counts[, lumA_idx] + 1))           # samples x genes
ens <- gsub("\\..*", "", colnames(lumA_expr))               # strip version
annot <- data.frame(
  probe         = colnames(lumA_expr),
  Gene.Symbol   = mapIds(org.Hs.eg.db, ens, "SYMBOL",  "ENSEMBL"),
  EntrezGene.ID = mapIds(org.Hs.eg.db, ens, "ENTREZID","ENSEMBL"),
  stringsAsFactors = FALSE)

# FOXC1 in the same sample order as lumA_expr rows
foxc1 <- as.numeric(foxc1_lumA)   # built from norm_counts[foxc1_id, lumA_idx]
stopifnot(length(foxc1) == nrow(lumA_expr))

# Compute each validated risk score
odx <- oncotypedx(data = lumA_expr, annot = annot, do.mapping = TRUE)$score
mp  <- gene70   (data = lumA_expr, annot = annot, do.mapping = TRUE)$score
ror <- rorS     (data = lumA_expr, annot = annot, do.mapping = TRUE)$score
# SCMGENE proliferation axis = AURKA expression (ESR1/ERBB2/AURKA model)
aurka <- as.numeric(lumA_expr[, annot$probe[which(annot$Gene.Symbol == "AURKA")[1]]])

tools <- list("Oncotype DX" = odx, "MammaPrint" = mp,
              "PAM50 ROR" = ror, "SCMGENE AURKA" = aurka)

cat("FOXC1 correlation with clinical genomic risk tools (LumA):\n")
for (nm in names(tools)) {
  ct <- cor.test(foxc1, tools[[nm]], method = "spearman", exact = FALSE)
  cat(sprintf("  %-16s rho=%+.3f  p=%.2e\n", nm, ct$estimate, ct$p.value))
}
```
```{r}
# ROR-S (subtype-correlation only) — exact, from genefu
ror_s <- rorS(data = lumA_expr, annot = annot, do.mapping = TRUE)$score

# PAM50 11-gene proliferation signature (Nielsen 2010 / Parker 2009)
prolif_genes <- c("BIRC5","CCNB1","CDC20","CEP55","MKI67","NDC80",
                  "NUF2","PTTG1","RRM2","TYMS","UBE2C")
pg <- annot$probe[annot$Gene.Symbol %in% prolif_genes]
prolif <- rowMeans(scale(lumA_expr[, pg]))

# ROR-P = subtype component + proliferation component
# Parker 2009 combines them with model-fit weights; z-scored sum below is a
# transparent stand-in. For the manuscript, use Parker's published proliferation
# coefficient so a reviewer reproduces your exact value.
ror_p <- as.numeric(scale(ror_s)) + as.numeric(scale(prolif))

ct <- function(x){ r <- cor.test(foxc1, x, method="spearman", exact=FALSE)
                   sprintf("rho=%+.3f  p=%.2e", r$estimate, r$p.value) }
cat("ROR-S (subtype only):    ", ct(ror_s),  "\n")
cat("Proliferation (PAM50):   ", ct(prolif), "\n")
cat("ROR-P (subtype+prolif):  ", ct(ror_p),  "\n")
```



#centaur symptoms unexpalinable by tumor purity
```{r}
# ============================================================
# Section 8.4 — Centaur is not explained by tumor purity / admixture
# ============================================================
library(genefu); library(survival); library(dplyr)
data(pam50.robust)   # load the PAM50 centroid model (not auto-loaded by library())

# ============================================================
# Part 1: FOXC1 vs PAM50 LumA-ness (pLumA) — the "by definition" lineage axis
# ============================================================
pam50_pred <- intrinsic.cluster.predict(
  sbt.model = pam50.robust, data = lumA_expr, annot = annot, do.mapping = TRUE)
print(colnames(pam50_pred$cor))            # confirm a "LumA" column exists
pLumA <- pam50_pred$cor[, "LumA"]
cat("FOXC1 vs pLumA: rho =",
    round(cor(foxc1, pLumA, method = "spearman"), 3),
    "(negative expected by construction)\n")

# ============================================================
# Part 2: FOXC1 vs purity in the CANONICAL radiation-naive cohort (surv_norad)
# ============================================================
# [REVIEW FIX] This chunk previously re-derived the radiation-naive cohort from brca.clin,
# producing a DIFFERENT n (182/24) vs the main pipeline's surv_norad (218/21). Now reuses
# surv_norad so the entire file reports one consistent cohort.
nr <- surv_norad
nr$centaur <- factor(ifelse(nr$group == "Centaur", "Centaur", "Non_centaur"),
                     levels = c("Non_centaur", "Centaur"))

# Align purity to surv_norad barcodes
.lumA_bc_all <- sample_meta_tumor$barcode[lumA_idx]
names(purity) <- .lumA_bc_all
nr$purity <- as.numeric(purity[nr$barcode])

cat("No-radiation cohort: n =", nrow(nr), ", events =", sum(nr$event), "\n")

# ---- FOXC1 vs purity (admixture check) ----
if (any(!is.na(nr$purity))) {
  cat("FOXC1 vs tumor purity: rho =",
      round(cor(nr$foxc1, nr$purity, method = "spearman",
                use = "complete.obs"), 3), "\n")
} else {
  cat("FOXC1 vs tumor purity: skipped (no real purity vector available).\n")
}

# ============================================================
# Part 3: purity-adjusted Cox in the canonical radiation-naive cohort
# ============================================================

if (any(!is.na(nr$purity))) {
  cox_p  <- coxph(Surv(time, event) ~ purity,           data = nr)
  cox_cp <- coxph(Surv(time, event) ~ centaur + purity, data = nr)
  cat("\n[REAL purity] alone HR:     ", round(exp(coef(cox_p)), 3), "\n")
  cat("[REAL purity] Centaur | purity HR: ",
      round(exp(coef(cox_cp))["centaurCentaur"], 3),
      "(does centaur add info beyond TRUE purity?)\n")
  print(summary(cox_cp)$coefficients)
} else {
  cat("\n[purity-adjusted Cox skipped] No real purity vector supplied.\n",
      " The luminal-basal 'lineage_contrast' is NOT a purity estimate and is not used here.\n")
}
```


---

# External Validation (scRNA-seq GSE176078)

```{r s9_load, eval=has_seurat}
sf <- params$seurat_object_path
scrna_ok <- file.exists(sf)
if (scrna_ok) { library(Seurat); seu <- readRDS(sf)
  ct_col <- "celltype_subset" }
```

## Normal Tissue: FOXC1 Marks Luminal Progenitors

```{r s9_normal, eval=has_seurat}
if (scrna_ok) {
  nl <- seu@meta.data[[ct_col]] %in% c("Luminal Progenitors","Mature Luminal")
  sn <- subset(seu, cells=colnames(seu)[nl])
  f1 <- FetchData(sn, vars="FOXC1")$FOXC1; g <- sn@meta.data[[ct_col]]
  lp_pct <- 100*mean(f1[g=="Luminal Progenitors"]>0)
  ml_pct <- 100*mean(f1[g=="Mature Luminal"]>0)
  pt <- prop.test(c(sum(f1[g=="Luminal Progenitors"]>0), sum(f1[g=="Mature Luminal"]>0)),
                   c(sum(g=="Luminal Progenitors"), sum(g=="Mature Luminal")))
  cat(sprintf("FOXC1+ in Luminal Progenitors: %.1f%%\n", lp_pct))
  cat(sprintf("FOXC1+ in Mature Luminal: %.1f%%\n", ml_pct))
  cat(sprintf("Fold enrichment: %.1fx, prop.test p = %.2e\n", lp_pct/ml_pct, pt$p.value))
} else cat("scRNA not loaded.\n")
```

## Hybrid vs Mosaic (Cancer LumA SC)

```{r s9_hybrid, eval=has_seurat}
if (scrna_ok) {
  sl <- subset(seu, cells=colnames(seu)[seu@meta.data[[ct_col]]=="Cancer LumA SC"])

  # Hybrid test 1: FOXC1-ESR1 co-detection enrichment
  cd <- FetchData(sl, vars=c("FOXC1","ESR1"))>0
  ft <- fisher.test(table(cd[,"FOXC1"], cd[,"ESR1"]))
  cat("FOXC1-ESR1 co-detection (hybrid test):\n")
  cat(sprintf("  Fisher OR = %.2f, p = %.2e (OR>1 = HYBRID, not admixture)\n",
      ft$estimate, ft$p.value))

  # === [REVIEW FIX] Depth-adjusted co-detection ============================
  # Shared per-cell library size co-drops BOTH genes (low-depth cells read 0/0, high-depth read
  # 1/1), which INFLATES a raw Fisher OR -- it does NOT bias it toward the null. So the manuscript's
  # "dropout biases toward null / conservative" sentence is the wrong direction. Control depth:
  .codet <- data.frame(
    ESR1_pos  = as.integer(cd[,"ESR1"]),
    FOXC1_pos = as.integer(cd[,"FOXC1"]),
    log_umi   = log1p(sl$nCount_RNA),    # sequencing depth
    log_gene  = log1p(sl$nFeature_RNA))  # genes detected (detection capacity)
  .m_adj <- glm(ESR1_pos ~ FOXC1_pos + log_umi + log_gene, data=.codet, family=binomial)
  cat(sprintf("  Depth-adjusted co-detection: OR = %.2f, p = %.2e (logistic, + log-UMI + log-#genes)\n",
      exp(coef(.m_adj)["FOXC1_pos"]), summary(.m_adj)$coefficients["FOXC1_pos","Pr(>|z|)"]))
  cat("  -> OR stays >1, p<0.05: depth-robust; rewrite the 'conservative/toward-null' sentence.\n")
  cat("  -> OR collapses to ~1: drop the Fisher OR; rely on within-patient rho + luminal retention.\n")
  # (If the assay is not named 'RNA', swap nCount_RNA / nFeature_RNA for the correct slot.)

  # Hybrid test 2: module scores, within-cell and within-patient correlation
  sl <- AddModuleScore(sl, features=list(c("ESR1","FOXA1","GATA3","XBP1","AGR2")), name="lum")
  sl <- AddModuleScore(sl, features=list(c("KRT5","KRT14","KRT17","FOXC1","EGFR")), name="bas")
  md <- sl@meta.data
  cat(sprintf("\nPooled luminal-basal score correlation: %.3f\n",
      cor(md$lum1, md$bas1, method="spearman")))
  pp <- md %>% group_by(orig.ident) %>% filter(n()>=30) %>%
    summarise(r = cor(lum1, bas1, method="spearman"), .groups="drop")
  cat(sprintf("Median WITHIN-patient correlation: %.3f (~0 = not mosaic)\n",
      median(pp$r, na.rm=TRUE)))

  # Hybrid test 2b: FOXC1+ cells RETAIN luminal identity (no loss of luminal program)
  fp <- FetchData(sl, vars="FOXC1")$FOXC1 > 0
  wl <- wilcox.test(md$lum1[fp], md$lum1[!fp])
  cat(sprintf("\nLuminal score, FOXC1+ vs FOXC1- cancer cells: %.3f vs %.3f (Wilcoxon p = %.2f)\n",
      mean(md$lum1[fp]), mean(md$lum1[!fp]), wl$p.value))
  cat("  -> FOXC1+ cells retain luminal identity (no significant loss)\n")

  # Hybrid test 3: basal score unimodality
  if (has_diptest) {
    dp <- diptest::dip.test(md$bas1)
    cat(sprintf("Basal score dip test p = %.3f (p>0.05 = unimodal = gradient)\n", dp$p.value))
  }

  # Hybrid test 4: identity retention by Wu's OWN SCSubtype classification
  #   Uses Wu's defined cell identities (SCSubtype calls), not hand-picked markers.
  #   Among cancer cells from the same ER+ tumors, do FOXC1+ cells stay Luminal-SC
  #   or flip to Basal-SC? Retained Luminal-SC = identity preserved per Wu.
  luma_pts   <- unique(sl@meta.data$orig.ident)                  # ER+ tumors (had LumA SC cells)
  cancer_lab <- grep("^Cancer.*SC$", as.character(unique(seu@meta.data[[ct_col]])), value=TRUE)
  in_set <- seu@meta.data$orig.ident %in% luma_pts &
            seu@meta.data[[ct_col]] %in% cancer_lab
  cc  <- subset(seu, cells = colnames(seu)[in_set])
  ccd <- FetchData(cc, vars="FOXC1")$FOXC1 > 0
  is_basal <- grepl("Basal", cc@meta.data[[ct_col]])
  cat("\nWu SCSubtype identity (cancer cells from ER+ tumors):\n")
  cat(sprintf("  Basal-SC call among FOXC1+ cells: %.1f%%\n", 100*mean(is_basal[ccd])))
  cat(sprintf("  Basal-SC call among FOXC1- cells: %.1f%%\n", 100*mean(is_basal[!ccd])))
  cat("  -> FOXC1+ cancer cells largely keep Wu's Luminal SCSubtype call = retained identity\n")
  # Gene-level alternative: load Wu's SCSubtype signature genes from the
  # Swarbricklab-code/BrCa_cell_atlas GitHub repo and AddModuleScore with those sets.
}
```

## Progenitor Markers in FOXC1+ Cancer Cells (honest: trend, NS)

```{r s9_prog_cancer, eval=has_seurat}
if (scrna_ok) {
  lp_mk <- c("KRT15","SLPI","ALDH1A3","PROM1","ELF5")
  lp_mk <- lp_mk[lp_mk %in% rownames(sl)]
  d <- FetchData(sl, vars=c("FOXC1", lp_mk)); d$fp <- d$FOXC1>0
  res <- data.frame()
  for (g in lp_mk) {
    wt <- wilcox.test(d[[g]][d$fp], d[[g]][!d$fp])
    res <- rbind(res, data.frame(marker=g,
      FOXC1pos=round(mean(d[[g]][d$fp]),4), FOXC1neg=round(mean(d[[g]][!d$fp]),4),
      ratio=round(mean(d[[g]][d$fp])/mean(d[[g]][!d$fp]),2), p=signif(wt$p.value,3)))
  }
  res$fdr <- signif(p.adjust(res$p, "BH"), 3)
  knitr::kable(res, caption="LP markers in FOXC1+ Cancer LumA cells (trend, mostly NS; BH-corrected)")
}
```

---

# Summary

```{r figure_export, message=FALSE, warning=FALSE}
# ============================================================
#  Assemble multi-panel figures and export to figures_final/
#  Composites use patchwork; panels are auto-labelled A, B, C...
#  Edit the groupings below freely -- each `fig*` block is one
#  manuscript figure. pheatmap (heatmap) and survminer (KM)
#  are saved via a graphics device since they are not ggplots.
# ============================================================
library(patchwork)
dir.create("figures_final", showWarnings = FALSE)

# cairo_pdf embeds Unicode (rho) and fonts cleanly; fall back to pdf if absent
.dev <- if (capabilities("cairo")) cairo_pdf else grDevices::pdf
save_fig <- function(p, file, w, h)
  ggsave(file.path("figures_final", file), plot = p, width = w, height = h, device = .dev)

# ---- Figure 1: defining the phenotype (distribution + FOXCUT-FOXC1) ----
fig1 <- (p1 | p2) + plot_annotation(tag_levels = "A")
save_fig(fig1, "fig1_phenotype_definition.pdf", 12, 5)

# ---- Figure 3: immune + stromal microenvironment (xCell2 panel if available) ----
if (exists("fig_xcell2") && !is.null(fig_xcell2)) {
  fig3 <- (fig_immune | fig_stromal) / fig_xcell2 + plot_annotation(tag_levels = "A")
  save_fig(fig3, "fig3_immune_stromal.pdf", 14, 13)
} else {
  fig3 <- (fig_immune | fig_stromal) + plot_annotation(tag_levels = "A")
  save_fig(fig3, "fig3_immune_stromal.pdf", 14, 8)
}

# ---- Figure 5: lineage quantification + proliferative quiescence ----
if (exists("fig_ssgsea") && !is.null(fig_ssgsea)) {
  fig5 <- fig_ssgsea / (fig_prolif_box | fig_prolif_cor) + plot_annotation(tag_levels = "A")
  save_fig(fig5, "fig5_lineage_proliferation.pdf", 12, 11)
} else {
  fig5 <- (fig_prolif_box | fig_prolif_cor) + plot_annotation(tag_levels = "A")
  save_fig(fig5, "fig5_lineage_proliferation.pdf", 10, 6)
}

# ---- Figure 6: mechanism -- progenitor / paracrine / inflammatory ----
fig6 <- fig_prog / fig_para / fig_inflam + plot_annotation(tag_levels = "A")
save_fig(fig6, "fig6_mechanism.pdf", 10, 15)

# ---- Figure 7: independence -- age confound + orthogonality to risk tools ----
fig7 <- (fig_foxc1_age | fig_age_partial) / fig_ortho + plot_annotation(tag_levels = "A")
save_fig(fig7, "fig7_independence.pdf", 12, 12)

# ---- Figure 4: dual-lineage heatmap (pheatmap gtable -> ggsave) ----
if (exists("ph_lineage")) {
  ggsave("figures_final/fig4_dual_lineage_heatmap.pdf", ph_lineage$gtable,
         width = 12, height = 8, device = .dev)
}

# ---- Figure 2: survival KM for FOXCUT and FOXC1, naive vs irradiated ----
# arrange_ggsurvplots() saves blank under ggplot2 >= 4.0 (S7); assemble $plot/$table via an
# explicit design grid (nesting + plot_layout locks dims and miscounts panels).
if (exists("km_foxcut_norad")) {
  km_foxcut_norad$plot <- km_foxcut_norad$plot + labs(tag = "A")
  km_foxc1_norad$plot  <- km_foxc1_norad$plot  + labs(tag = "B")
  km_foxcut_rad$plot   <- km_foxcut_rad$plot   + labs(tag = "C")
  km_foxc1_rad$plot    <- km_foxc1_rad$plot    + labs(tag = "D")
  km_design <- "AB\nCD\nEF\nGH"   # rows: norad curves / norad tables / rad curves / rad tables
  km_fig <-
    km_foxcut_norad$plot  + km_foxc1_norad$plot +
    km_foxcut_norad$table + km_foxc1_norad$table +
    km_foxcut_rad$plot    + km_foxc1_rad$plot +
    km_foxcut_rad$table   + km_foxc1_rad$table +
    plot_layout(design = km_design, heights = c(3, 1, 3, 1))
  ggsave("figures_final/fig2_survival_km.pdf", km_fig, width = 14, height = 13, device = .dev)
}

cat("Saved to figures_final/:\n"); print(list.files("figures_final"))
```

```{r summary, echo=FALSE}
cat("CENTAUR STATE — VALIDATED FINDINGS\n")
cat("==================================\n\n")
cat("S1: FOXC1 normal distribution in LumA; median split\n")
cat("S2: Survival HR~0.28 (no-rad); censor-robust; PH holds\n")
cat("S3-5: Immune/stromal/xCell2 correlations\n")
cat("S6: Dual lineage; proliferative quiescence; progenitor; paracrine by source; 3-tier inflammatory\n")
cat("S7: Progenitor mediates ~67%; cytotoxic dominant; age=suppressor\n")
cat("S8: Orthogonal to Oncotype/MammaPrint/ROR; cuts across Poudel\n")
cat("S9: FOXC1 marks normal progenitors (3.8x); hybrid>mosaic; cancer-cell progenitor trend NS\n")
```

---

*Generated `r Sys.time()` | Median split log2=`r CENTAUR_LOG2` | No-rad n=`r nrow(surv_norad)`, `r sum(surv_norad$event)` events*

---

# Review Additions — optional mediation & validation scope

## MHC class I and xCell2-CD8 mediation (the two adjustments named in the abstract)

```{r add_mhc_cd8_mediation}
# [ADDED, runs] The two extra mediation adjustments named in the abstract: an MHC class I
# antigen-presentation score and an xCell2-derived CD8+ T-cell score, in the same age+stage
# adjusted style as s7_chain. CAVEAT (unchanged and important): adding a covariate collinear
# with the Centaur axis attenuates the HR mechanically -- this is "consistent with", NOT proof
# of mediation or a "causal chain / primary mechanism". Report it that way.
med_rows <- list()

# MHC class I antigen-presentation score
surv_cc$mhc1_z <- make_score(c("HLA-A","HLA-B","HLA-C","B2M","TAP1","TAP2"), surv_cc$barcode)
med_rows[["+ MHC class I"]] <- extract_hr(
  coxph(Surv(time,event) ~ group+age+stage_clean+mhc1_z, data=surv_cc))

# xCell2-derived CD8+ T-cell score (active CD8, matching paper Section J)
xf <- "results/rds/xcell2_scores.rds"
if (file.exists(xf)) {
  xc <- readRDS(xf)
  if (any(grepl("^TCGA", rownames(xc))) && !any(grepl("^TCGA", colnames(xc)))) xc <- t(xc)
  cd8_all <- grep("CD8", rownames(xc), value=TRUE, ignore.case=TRUE)
  cd8_row <- cd8_all[grepl("CD8-positive", cd8_all, ignore.case=TRUE) &
                     !grepl("naive|memory|exhaust", cd8_all, ignore.case=TRUE)][1]
  if (is.na(cd8_row) && length(cd8_all)) cd8_row <- cd8_all[1]
  if (!is.na(cd8_row)) {
    nb <- function(x) toupper(gsub("\\.", "-", x))
    xb <- nb(colnames(xc)); sb <- nb(surv_cc$barcode); k <- NA
    for (kk in c(28,19,16,15,12)) if (length(intersect(substr(sb,1,kk), substr(xb,1,kk)))>0) { k <- kk; break }
    if (!is.na(k)) {
      sc <- as.numeric(xc[cd8_row, ]); names(sc) <- substr(xb,1,k); sc <- sc[!duplicated(names(sc))]
      surv_cc$cd8x_z <- scale(sc[substr(sb,1,k)])[,1]
      sc8 <- surv_cc[!is.na(surv_cc$cd8x_z), ]
      med_rows[["+ xCell2 CD8"]] <- extract_hr(
        coxph(Surv(time,event) ~ group+age+stage_clean+cd8x_z, data=sc8))
    }
  }
}

out <- do.call(rbind, c(
  list(cbind(Model="age+stage baseline",
             extract_hr(coxph(Surv(time,event)~group+age+stage_clean, data=surv_cc)))),
  lapply(names(med_rows), function(n) cbind(Model=n, med_rows[[n]]))))
knitr::kable(out, row.names=FALSE,
  caption="Abstract-named mediators (MHC class I, xCell2 CD8); HR attenuation is collinearity, not proof of mediation")
if (is.null(surv_cc$cd8x_z)) cat("xCell2 CD8 step skipped (scores file or CD8 row not found).\n")
```
---

# Review Additions — paper analyses consolidated from the other pipelines

These blocks were present in the figure/mechanism pipelines and correspond to claims made in the
manuscript, but were not in the master. Each is adapted to this file's objects (`norm_counts`,
`lumA_idx`, `foxc1_lumA`, `get_ids`/`lookup_id`, the `imm` table, the new GSVA `ssgseaParam` API)
and guarded so the knit degrades gracefully when optional inputs/packages are unavailable.

## [backs Section J] xCell2 score provenance

```{r add_xcell2_provenance}
cat("xCell2 cache path: results/rds/xcell2_scores.rds\n")
cat("If absent, this report attempts to rebuild it in the earlier reproducible_input_cache chunk using results/rds/tpm_counts_tumor.rds.\n")
cat("xCell2 package available:", requireNamespace("xCell2", quietly = TRUE), "\n")
cat("xCell2 scores available:", file.exists("results/rds/xcell2_scores.rds"), "\n")
```

## [backs Section H.5] Chemokine–myeloid dissociation

```{r add_chemokine_myeloid_dissociation}
# Recruitment chemokines track FOXC1, but the myeloid CELL markers that would receive them do not
# -> "a recruitment signal without the receiving population". Uses the Section 3 'imm' table.
if (exists("imm")) {
  diss_genes <- c("CCL19","CCL21","CXCL13","CXCL1","CCL2","CCL4","CXCL8",   # recruitment chemokines
                  "CD68","ITGAM","CD163","MRC1","CSF1R")                     # myeloid-cell markers
  diss <- imm[as.character(imm$gene) %in% diss_genes, c("gene","cat","rho","p","fdr")]
  diss$gene <- as.character(diss$gene)
  knitr::kable(diss[order(-diss$rho), ], row.names = FALSE,
    caption = "Chemokine-myeloid dissociation: recruitment chemokines (rho>0) vs myeloid-cell markers (rho~0)")
} else {
  cat("Run Section 3 first (the 'imm' table is built there).\n")
}
```

## [backs Section D/K: "not EMT — epithelial identity retained"] Epithelial vs EMT markers

```{r add_emt_epithelial_check}
# Expectation: epithelial markers are retained (not strongly negative with FOXC1) and EMT markers
# are NOT elevated -> the basal program is a lineage hybrid, not an epithelial-mesenchymal transition.
epi_markers <- c("EPCAM","CDH1","KRT18","KRT19","KRT8","MUC1")
emt_markers <- c("VIM","CDH2","SNAI1","SNAI2","TWIST1","ZEB1","ZEB2","FN1")
.emt_one <- function(sym, cat) {
  gid <- lookup_id(sym); if (is.na(gid)) return(NULL)
  ct <- suppressWarnings(cor.test(foxc1_lumA, as.numeric(norm_counts[gid, lumA_idx]),
                                  method = "spearman", exact = FALSE))
  data.frame(gene = sym, category = cat, rho = round(ct$estimate, 3), p = signif(ct$p.value, 3))
}
emt_res <- do.call(rbind, c(lapply(epi_markers, .emt_one, cat = "Epithelial"),
                            lapply(emt_markers, .emt_one, cat = "EMT")))
emt_res$fdr <- signif(p.adjust(emt_res$p, "BH"), 3)
knitr::kable(emt_res[order(emt_res$category, -emt_res$rho), ], row.names = FALSE,
  caption = "FOXC1 vs epithelial / EMT markers in LumA (BH-corrected) -- tests the 'not EMT' claim")
```

## [backs Limitations & Section K.2: compositional / CAF confound] CAF-adjusted partial correlation

```{r add_caf_confound_partialcor, eval=has_gsva}
# Does adjusting for CAF/FRC content kill the FOXC1 -> chemokine correlation?
#   drop_pct < 20%  -> tumor-autonomous; 20-50% -> partially CAF-mediated; >50% -> largely CAF-mediated.
# This is the confound a reviewer will specifically demand. (From chemokine_mechanism_alternatives.)
caf_sigs <- list(
  CAF_general = c("FAP","ACTA2","PDGFRA","PDGFRB","S100A4","COL1A1","COL1A2","COL3A1","DCN","LUM"),
  FRC_like    = c("PDPN","CCL19","CCL21","IL33","CXCL12","CD34","BMP4","ICAM1","VCAM1"),
  iCAF        = c("IL6","CXCL12","CXCL1","CXCL2","HAS1","PDGFRA","DPT"),
  myCAF       = c("ACTA2","MYL9","TAGLN","TPM1","TPM2","MMP11","POSTN"),
  HEV         = c("PLVAP","GLYCAM1","CCL21","CHST4","MADCAM1","SELE","SELP","ICAM1"))
caf_sig_ids <- lapply(caf_sigs, get_ids)
caf_f <- "results/rds/stromal_signatures.rds"
if (file.exists(caf_f)) {
  caf_ss <- readRDS(caf_f)
} else {
  par_caf <- ssgseaParam(exprData = as.matrix(log2(norm_counts + 1)), geneSets = caf_sig_ids)
  caf_ss  <- gsva(par_caf, verbose = FALSE); saveRDS(caf_ss, caf_f)
}
caf_cols <- match(sample_meta_tumor$barcode[lumA_idx], colnames(caf_ss))
ok <- !is.na(caf_cols); fx_ok <- foxc1_lumA[ok]; caf_ok <- caf_cols[ok]

sig_cor <- do.call(rbind, lapply(rownames(caf_ss), function(s) {
  ct <- cor.test(fx_ok, as.numeric(caf_ss[s, caf_ok]), method = "spearman", exact = FALSE)
  data.frame(signature = s, rho = round(ct$estimate, 3), p = signif(ct$p.value, 3)) }))
knitr::kable(sig_cor, row.names = FALSE, caption = "FOXC1 vs stromal/CAF ssGSEA signatures in LumA")

if (requireNamespace("ppcor", quietly = TRUE)) {
  .pcor_one <- function(chemok, ctrl_sig) {
    gid <- lookup_id(chemok); if (is.na(gid)) return(NULL)
    tv <- as.numeric(norm_counts[gid, lumA_idx])[ok]; cv <- as.numeric(caf_ss[ctrl_sig, caf_ok])
    d  <- data.frame(fc1 = fx_ok, target = tv, ctrl = cv); d <- d[complete.cases(d), ]
    raw <- cor(d$fc1, d$target, method = "spearman")
    pc  <- ppcor::pcor.test(d$fc1, d$target, d$ctrl, method = "spearman")
    data.frame(chemokine = chemok, control = ctrl_sig, raw_rho = round(raw, 3),
               partial_rho = round(pc$estimate, 3), partial_p = signif(pc$p.value, 3),
               drop_pct = round(100 * (abs(raw) - abs(pc$estimate)) / abs(raw), 1)) }
  pcr <- do.call(rbind, list(
    .pcor_one("CCL19","FRC_like"),    .pcor_one("CCL21","FRC_like"),
    .pcor_one("CCL5","FRC_like"),     .pcor_one("CXCL9","FRC_like"),
    .pcor_one("CCL19","CAF_general"), .pcor_one("CCL21","CAF_general"),
    .pcor_one("CCL5","CAF_general"),  .pcor_one("CCL21","HEV")))
  knitr::kable(pcr, row.names = FALSE,
    caption = "Partial correlation: FOXC1-chemokine controlling for stromal signature (drop_pct = % attenuation)")
} else {
  cat("Install 'ppcor' to run the CAF-confound partial-correlation test.\n")
}
```

## Exploratory mechanism tests — NOT in the manuscript (retained for completeness)

These test mechanisms the paper does **not** claim (FOXC1-direct transcription; hypoxia upstream).
Kept eval=FALSE and demarcated so they are not mistaken for reported results.

```{r add_exploratory_hypoxia, eval=FALSE}
# Hypoxia (Buffa 2010) as an alternative upstream driver of the chemokine signal.
buffa <- c("VEGFA","SLC2A1","CA9","LDHA","PGK1","HK2","ALDOA","ENO1","NDRG1","BNIP3",
           "P4HA1","ADM","MIF","TPI1","PGAM1")
par_hyp <- ssgseaParam(exprData = as.matrix(log2(norm_counts + 1)),
                       geneSets = list(hypoxia = get_ids(buffa)))
hyp_ss  <- gsva(par_hyp, verbose = FALSE)
hcols   <- match(sample_meta_tumor$barcode[lumA_idx], colnames(hyp_ss))
hok     <- !is.na(hcols)
if (requireNamespace("ppcor", quietly = TRUE)) {
  d <- data.frame(fc1 = foxc1_lumA[hok],
                  ccl19 = as.numeric(norm_counts[lookup_id("CCL19"), lumA_idx])[hok],
                  hyp   = as.numeric(hyp_ss["hypoxia", hcols[hok]]))
  d <- d[complete.cases(d), ]
  print(ppcor::pcor.test(d$fc1, d$ccl19, d$hyp, method = "spearman"))  # does hypoxia explain it?
}
```

### Optional Test 3: FOXC1 motif enrichment in chemokine promoters

This exploratory test asks whether chemokine promoters contain predicted FOXC1 binding motifs more often than a background set of expressed genes. It is included for repository completeness, but it is **disabled by default** because it requires JASPAR, genome, and promoter-annotation packages. It does not prove direct FOXC1 regulation; it only tests motif compatibility.

```{r exploratory_foxc1_motif_promoters, eval=params$run_foxc1_motif_test}
# Optional heavy-dependency promoter-motif scan.
# Set params$run_foxc1_motif_test = TRUE to run.

motif_needed <- c(
  "JASPAR2020", "TFBSTools", "BSgenome", "BSgenome.Hsapiens.UCSC.hg38",
  "TxDb.Hsapiens.UCSC.hg38.knownGene", "GenomicFeatures", "GenomicRanges",
  "GenomeInfoDb", "Biostrings", "org.Hs.eg.db", "AnnotationDbi"
)
missing_motif <- motif_needed[!vapply(motif_needed, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_motif) > 0 && isTRUE(params$install_missing_packages)) {
  ensure_packages(missing_motif, bioc = TRUE)
  missing_motif <- motif_needed[!vapply(motif_needed, requireNamespace, logical(1), quietly = TRUE)]
}

if (length(missing_motif) > 0) {
  cat("Skipping FOXC1 promoter motif scan; missing package(s): ",
      paste(missing_motif, collapse = ", "),
      "\nInstall these packages or knit with params$install_missing_packages = TRUE.\n", sep = "")
} else {
  suppressPackageStartupMessages({
    library(JASPAR2020)
    library(TFBSTools)
    library(BSgenome.Hsapiens.UCSC.hg38)
    library(TxDb.Hsapiens.UCSC.hg38.knownGene)
    library(GenomicFeatures)
    library(GenomicRanges)
    library(GenomeInfoDb)
    library(Biostrings)
  })

  dir.create("results/tables", recursive = TRUE, showWarnings = FALSE)
  motif_scan_file <- "results/tables/foxc1_motif_promoter_scan.csv"
  motif_enrich_file <- "results/tables/foxc1_motif_promoter_enrichment.csv"

  get_jaspar_object <- function() {
    if (exists("JASPAR2020", envir = asNamespace("JASPAR2020"), inherits = FALSE)) {
      get("JASPAR2020", envir = asNamespace("JASPAR2020"))
    } else {
      JASPAR2020::JASPAR2020
    }
  }

  jaspar_db <- get_jaspar_object()
  pfms <- TFBSTools::getMatrixSet(
    jaspar_db,
    opts = list(species = 9606, all_versions = FALSE, collection = "CORE", name = "FOXC1")
  )
  if (length(pfms) == 0) stop("No FOXC1 motif found in JASPAR2020 with the selected options.")
  pfm_names <- vapply(pfms, TFBSTools::name, character(1))
  pfm_idx <- grep("FOXC1", pfm_names, ignore.case = TRUE)[1]
  if (is.na(pfm_idx)) pfm_idx <- 1
  foxc1_pwm <- TFBSTools::toPWM(pfms[[pfm_idx]], type = "prob")
  cat("Using JASPAR motif:", TFBSTools::name(pfms[[pfm_idx]]), "\n")

  txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene::TxDb.Hsapiens.UCSC.hg38.knownGene
  genome <- BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38

  promoter_seqs_for_symbols <- function(symbols) {
    symbols <- unique(symbols[!is.na(symbols) & nzchar(symbols)])
    if (length(symbols) == 0) return(Biostrings::DNAStringSet())
    entrez <- AnnotationDbi::mapIds(
      org.Hs.eg.db, keys = symbols, column = "ENTREZID", keytype = "SYMBOL", multiVals = "first"
    )
    map <- data.frame(symbol = names(entrez), entrez = as.character(entrez), stringsAsFactors = FALSE)
    map <- map[!is.na(map$entrez) & !duplicated(map$entrez), , drop = FALSE]
    if (nrow(map) == 0) return(Biostrings::DNAStringSet())

    g <- GenomicFeatures::genes(txdb)
    map <- map[map$entrez %in% names(g), , drop = FALSE]
    if (nrow(map) == 0) return(Biostrings::DNAStringSet())

    g <- g[map$entrez]
    pr <- GenomicRanges::promoters(
      g,
      upstream = as.integer(params$promoter_upstream_bp),
      downstream = as.integer(params$promoter_downstream_bp)
    )
    common_seq <- intersect(GenomeInfoDb::seqlevels(pr), GenomeInfoDb::seqlevels(genome))
    pr <- GenomeInfoDb::keepSeqlevels(pr, common_seq, pruning.mode = "coarse")
    GenomeInfoDb::seqinfo(pr) <- GenomeInfoDb::seqinfo(genome)[GenomeInfoDb::seqlevels(pr)]
    pr <- GenomicRanges::trim(pr)
    if (length(pr) == 0) return(Biostrings::DNAStringSet())

    seqs <- BSgenome::getSeq(genome, pr)
    symbol_by_entrez <- setNames(map$symbol, map$entrez)
    names(seqs) <- unname(symbol_by_entrez[names(pr)])
    seqs[!duplicated(names(seqs))]
  }

  chemokine_sets <- list(
    TLS_lymphoid = c("CCL19", "CCL21", "CXCL13"),
    T_cell_recruiting = c("CXCL9", "CXCL10", "CCL5", "CCL4"),
    Myeloid_recruiting = c("CCL2", "CXCL1", "CXCL8", "CXCL12")
  )

  target_symbols <- unique(unlist(chemokine_sets))
  all_symbols <- unique(na.omit(unname(gene_id_map)))
  all_symbols <- setdiff(all_symbols, target_symbols)
  set.seed(1)
  bg_n <- min(as.integer(params$motif_background_genes), length(all_symbols))
  background_symbols <- sample(all_symbols, bg_n)

  scan_design <- dplyr::bind_rows(
    lapply(names(chemokine_sets), function(nm) {
      data.frame(set = nm, symbol = chemokine_sets[[nm]], stringsAsFactors = FALSE)
    }),
    data.frame(set = "background_expressed", symbol = background_symbols, stringsAsFactors = FALSE)
  )

  promoter_seqs <- promoter_seqs_for_symbols(unique(scan_design$symbol))
  scan_design <- scan_design[scan_design$symbol %in% names(promoter_seqs), , drop = FALSE]

  site_counts <- vapply(names(promoter_seqs), function(sym) {
    sites <- TFBSTools::searchSeq(
      foxc1_pwm,
      promoter_seqs[[sym]],
      min.score = params$foxc1_motif_min_score,
      strand = "*"
    )
    length(sites)
  }, integer(1))

  scan_design$foxc1_motif_count <- unname(site_counts[scan_design$symbol])
  scan_design$foxc1_motif_present <- scan_design$foxc1_motif_count > 0
  readr::write_csv(scan_design, motif_scan_file)

  enrich <- dplyr::bind_rows(lapply(setdiff(unique(scan_design$set), "background_expressed"), function(st) {
    a <- sum(scan_design$set == st & scan_design$foxc1_motif_present)
    b <- sum(scan_design$set == st & !scan_design$foxc1_motif_present)
    c <- sum(scan_design$set == "background_expressed" & scan_design$foxc1_motif_present)
    d <- sum(scan_design$set == "background_expressed" & !scan_design$foxc1_motif_present)
    ft <- fisher.test(matrix(c(a, b, c, d), nrow = 2, byrow = TRUE))
    data.frame(
      chemokine_set = st,
      target_genes_scanned = a + b,
      target_motif_positive = a,
      background_genes_scanned = c + d,
      background_motif_positive = c,
      odds_ratio = unname(ft$estimate),
      p_value = ft$p.value,
      stringsAsFactors = FALSE
    )
  }))
  enrich$fdr <- p.adjust(enrich$p_value, method = "BH")
  readr::write_csv(enrich, motif_enrich_file)

  knitr::kable(scan_design %>% dplyr::filter(set != "background_expressed"),
               caption = "FOXC1 motif scan in selected chemokine promoters")
  knitr::kable(enrich,
               digits = 3,
               caption = "Exploratory FOXC1 motif enrichment in chemokine promoters vs expressed-gene background")
  cat("Saved motif scan:", motif_scan_file, "\n")
  cat("Saved enrichment table:", motif_enrich_file, "\n")
}
```

### Optional Test 4: single-cell FOXC1/CCL19 co-expression by cell type

This exploratory test asks whether \textit{FOXC1} and \textit{CCL19} are co-detected in the same single cells, stratified by Wu et al. cell-type annotations when a processed Seurat object is available. It is disabled by default because it requires `params$seurat_object_path` to point to a processed Seurat object.

```{r exploratory_scrna_foxc1_ccl19_by_celltype, eval=params$run_scrna_ccl19_test}
# Optional single-cell FOXC1/CCL19 co-expression scan.
# Set params$run_scrna_ccl19_test = TRUE and params$seurat_object_path to a processed Seurat object.

sc_needed <- c("Seurat", "dplyr", "ggplot2", "readr")
missing_sc <- sc_needed[!vapply(sc_needed, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_sc) > 0 && isTRUE(params$install_missing_packages)) {
  ensure_packages(missing_sc, bioc = FALSE)
  missing_sc <- sc_needed[!vapply(sc_needed, requireNamespace, logical(1), quietly = TRUE)]
}

if (length(missing_sc) > 0) {
  cat("Skipping single-cell FOXC1/CCL19 test; missing package(s): ",
      paste(missing_sc, collapse = ", "),
      "\nInstall these packages or knit with params$install_missing_packages = TRUE.\n", sep = "")
} else if (!file.exists(params$seurat_object_path)) {
  cat("Skipping single-cell FOXC1/CCL19 test; Seurat object not found at: ",
      params$seurat_object_path, "\n", sep = "")
} else {
  suppressPackageStartupMessages({
    library(Seurat)
    library(dplyr)
    library(ggplot2)
  })

  dir.create("results/tables", recursive = TRUE, showWarnings = FALSE)
  dir.create("results/figures", recursive = TRUE, showWarnings = FALSE)

  sc_obj <- readRDS(params$seurat_object_path)
  md <- sc_obj@meta.data
  ct_candidates <- c(
    "celltype_subset", "cell_type", "celltype", "CellType", "annotation",
    "major_celltype", "SCSubtype", "subtype", "cell_type_major"
  )
  ct_col <- ct_candidates[ct_candidates %in% colnames(md)][1]
  if (is.na(ct_col)) {
    stop("Could not identify a cell-type metadata column. Add one of: ",
         paste(ct_candidates, collapse = ", "), call. = FALSE)
  }
  if (!all(c("FOXC1", "CCL19") %in% rownames(sc_obj))) {
    missing_genes <- setdiff(c("FOXC1", "CCL19"), rownames(sc_obj))
    cat("Skipping single-cell FOXC1/CCL19 test; missing gene(s) in Seurat object: ",
        paste(missing_genes, collapse = ", "), "\n", sep = "")
  } else {
    fd <- Seurat::FetchData(sc_obj, vars = c("FOXC1", "CCL19"))
    fd$cell_type <- as.character(md[[ct_col]])
    fd$FOXC1_pos <- fd$FOXC1 > 0
    fd$CCL19_pos <- fd$CCL19 > 0

    fisher_one <- function(d) {
      tab <- table(
        factor(d$FOXC1_pos, levels = c(FALSE, TRUE)),
        factor(d$CCL19_pos, levels = c(FALSE, TRUE))
      )
      if (sum(tab) < 20 || any(rowSums(tab) == 0) || any(colSums(tab) == 0)) {
        return(c(odds_ratio = NA_real_, p_value = NA_real_))
      }
      ft <- fisher.test(tab)
      c(odds_ratio = unname(ft$estimate), p_value = ft$p.value)
    }

    coexpr_by_type <- dplyr::bind_rows(lapply(split(fd, fd$cell_type), function(d) {
      ft <- fisher_one(d)
      data.frame(
        cell_type = unique(d$cell_type)[1],
        n_cells = nrow(d),
        foxc1_detected_pct = 100 * mean(d$FOXC1_pos),
        ccl19_detected_pct = 100 * mean(d$CCL19_pos),
        co_detected_pct = 100 * mean(d$FOXC1_pos & d$CCL19_pos),
        ccl19_among_foxc1_pos_pct = ifelse(sum(d$FOXC1_pos) > 0,
                                           100 * mean(d$CCL19_pos[d$FOXC1_pos]), NA_real_),
        fisher_or = ft["odds_ratio"],
        fisher_p = ft["p_value"],
        stringsAsFactors = FALSE
      )
    })) %>%
      dplyr::arrange(dplyr::desc(co_detected_pct), dplyr::desc(n_cells))
    coexpr_by_type$fdr <- p.adjust(coexpr_by_type$fisher_p, method = "BH")

    out_csv <- "results/tables/scrna_foxc1_ccl19_coexpression_by_cell_type.csv"
    readr::write_csv(coexpr_by_type, out_csv)

    p_co <- coexpr_by_type %>%
      dplyr::filter(n_cells >= 20) %>%
      ggplot(aes(co_detected_pct, reorder(cell_type, co_detected_pct))) +
      geom_col() +
      labs(
        x = "FOXC1/CCL19 co-detected cells (%)",
        y = NULL,
        title = "Single-cell FOXC1/CCL19 co-detection by cell type"
      ) +
      theme_minimal(base_size = 12)
    print(p_co)
    ggsave("results/figures/scrna_foxc1_ccl19_coexpression_by_cell_type.pdf",
           p_co, width = 8, height = 6)

    knitr::kable(coexpr_by_type,
                 digits = 3,
                 caption = "Exploratory single-cell FOXC1/CCL19 co-expression by cell type")
    cat("Saved single-cell co-expression table:", out_csv, "\n")
    cat("Interpretation note: co-detection supports cellular-source plausibility only; it does not prove direct FOXC1 regulation of CCL19.\n")
  }
}
```

---

# Session Information

```{r session_info}
sessionInfo()
writeLines(capture.output(sessionInfo()), "session_info.txt")
```

# Code and Data Availability Template

```{r code_availability_text, results='asis'}
cat(sprintf("Code used to generate this analysis report and the manuscript figures is available at: %s\n\n", repo_url))
cat("Large public datasets are not committed directly to the repository. The R Markdown report rebuilds TCGA-BRCA-derived expression, clinical, subtype, tumor-purity, and xCell2-derived objects when cached files are absent and the corresponding parameters are enabled. The proposed FOXCUT/FOXC1-centered model is supported by observational bulk and single-cell analyses, but requires independent spatial, longitudinal, treatment-annotated, or perturbational validation.\n")
```
